---
title: "Cleaning"
format: html
---

## Goals of this notebook

1. Import all data
2. Merge all sheets
3. Remove duplicate rows
4. Geocode
5. Clean names in OpenRefine
6. Fix date data types and group by dates
7. Create and rename final columns
8. Export clean data

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(stringr)
library(readxl)
```

## 1. Importing

Data set 1.

```{r}
unhoused_raw_1 <- suppressWarnings(read_excel("data-raw/unhoused-raw-1.xlsx"))
```

Data set 2.

```{r}
unhoused_raw_2 <- suppressWarnings(read_excel("data-raw/unhoused-raw-2.xlsx"))
```

Data set 3.

```{r}
unhoused_raw_3 <- suppressWarnings(read_excel("data-raw/unhoused-raw-3.xlsx"))
```

Data set 4.

```{r}
unhoused_raw_4 <- suppressWarnings(read_excel("data-raw/unhoused-raw-4.xlsx"))
```

Data set 5.

```{r}
unhoused_raw_5 <- suppressWarnings(read_excel("data-raw/unhoused-raw-5.xlsx"))
```

## 2. Merging data sets

### 2.1 Evaluating merge readiness

Ensuring column names match.

```{r}
unhoused_raw_1 |> glimpse()
unhoused_raw_2 |> glimpse()
unhoused_raw_3 |> glimpse()
unhoused_raw_4 |> glimpse()
unhoused_raw_5 |> glimpse()
```

### 2.3 Cleaning names

#### 2.3.1 Data set 1

```{r}
unhoused_names_1 <- unhoused_raw_1 |> 
  rename(defendant_address = Defendent_Address) |> # this was misspelled
  clean_names()

unhoused_names_1 |> glimpse()
```

#### 2.3.2 Data set 2

```{r}
unhoused_names_2 <- unhoused_raw_2 |> 
  clean_names()

unhoused_names_2 |> glimpse()
```

#### 2.3.3 Data set 3

```{r}
unhoused_names_3 <- unhoused_raw_3 |> 
  clean_names()

unhoused_names_3 |> glimpse()
```

#### 2.3.4 Data set 4

```{r}
unhoused_names_4 <- unhoused_raw_4 |> 
  clean_names()

unhoused_names_4 |> glimpse()
```

#### 2.3.5 Data set 5

```{r}
unhoused_names_5 <- unhoused_raw_5 |> 
  clean_names()

unhoused_names_5 |> glimpse()
```

### 2.4 Preparing to merge

#### 2.4.1 Data set 1

Fixing phone_number data type, fixing misspellings and adding data_origin column to data set 1.

```{r}
unhoused_tidy_1 <- unhoused_names_1 |>
  mutate(phone_number = as.character(phone_number),
         data_origin = "unhoused_1"
         ) |> 
  rename(total_fine_amount_due = totla_fine_amount_due,
         total_fine_amount_paid = totla_fine_amount_paid,
         total_fine_amount_dismissed = totla_fine_amount_dismissed)

unhoused_tidy_1 |> glimpse()
```

#### 2.4.2 Data set 2

Adding data_origin column to data set 2 and renaming columns to match data set 1.

```{r}
unhoused_tidy_2 <- unhoused_names_2 |> 
  mutate(data_origin = "unhoused_2") |> 
  rename(phone_number = defendant_phone_number)

unhoused_tidy_2 |> glimpse()
```

#### 2.4.3 Data set 3

Adding data_origin column to data set 3 and renaming columns to match the other data sets.

```{r}
unhoused_tidy_3 <- unhoused_names_3 |> 
  mutate(data_origin = "unhoused_3") |> 
  rename(phone_number = defendant_phone_number)

unhoused_tidy_3 |> glimpse()
```

#### 2.4.4 Data set 4

Fixing phone_number data type and adding data_origin column to data set 4. Renaming columns to match the other data sets.

```{r}
unhoused_tidy_4 <- unhoused_names_4 |> 
  rename(phone_number = defendant_phone_number) |> 
  mutate(phone_number = as.character(phone_number),
         data_origin = "unhoused_4")

unhoused_tidy_4 |> glimpse()
```

#### 2.4.5 Data set 5

Fixing phone_number data type and adding data_origin column to data set 5. Renaming columns to match the other data sets.

```{r}
unhoused_tidy_5 <- unhoused_names_5 |> 
  rename(phone_number = defendant_phone_number) |> 
  mutate(phone_number = as.character(phone_number),
         data_origin = "unhoused_5")

unhoused_tidy_5 |> glimpse()
```

### 2.5 Merging sheets

Merging sheets.

```{r}
unhoused_merged <- bind_rows(unhoused_tidy_1, unhoused_tidy_2, unhoused_tidy_3, unhoused_tidy_4, unhoused_tidy_5)

unhoused_merged |> glimpse()
```

## 3. Remove duplicate rows

### 3.1 Finding dupes

This shows both duplicates, so there are actually 219 (438/2) pairs of duplicates.

```{r}
unhoused_merged |> 
   get_dupes(!c(data_origin, defendant_address))
```

### 3.2 Removing dupes

Proof it worked: subtract the number of duplicate pairs (219) from the total number of rows (31,626), that is how many rows are outputted below (31,407).

```{r}
unhoused_no_dupes <- unhoused_merged |>
  arrange(desc(data_origin)) |> # distinct() keeps the first row between two duplicates, so arranging it in descending order by data_origin, it makes sure we keep the most recent row
  distinct(across(!c(
    data_origin, 
    defendant_address)),
    .keep_all = TRUE)

unhoused_no_dupes |> glimpse()
```

## 4. Geocoding

### 4.1 Reformatting addresses

Reformatting addresses to prepare to geocode externally using geocodeio.

```{r}
unhoused_reformatted <- unhoused_no_dupes |>
  mutate(full_address = paste(str_remove(offense_location, ","),
                              "Houston, TX", sep = ", "))

unhoused_reformatted |> glimpse()
```

### 4.2 Geocoding sets 1-4

#### 4.2.1 Isolating sets 1-4

Isolating data sets from 1-4 to geocode.

```{r}
unhoused_to_geocode_1 <- unhoused_reformatted |> 
  filter(!c(data_origin == "unhoused_5"))

unhoused_to_geocode_1 |> glimpse()
```

#### 4.2.2 Exporting pre-geocode sets 1-4

Exporting data sets 1-4 to geocode in geocodio.

```{r}
unhoused_to_geocode_1 |>
  write_csv("data-processed/unhoused-to-geocode-1.csv")
```

#### 4.2.3 Importing geocoded

Importing the latest batch of geocoded data from geocodio and cleaning column names. @NOTE: There are more rows here than previously exported because this batch was geocoded before receiving fifth data set. unhoused_4 decreased in rows after introducing unhoused_5 because there was overlap between the two, and we kept most updated data in unhoused_5 --- so no rows were lost, just transferred. This just means it will be necessary to remove duplicates again after merging in 4.4.

```{r}
unhoused_geocoded_1 <- read_csv("data-processed/unhoused-geocoded-1.csv") |> 
  clean_names() # the new geocoded columns didn't have clean names

unhoused_geocoded_1 |> glimpse()
```

### 4.3 Geocoding set 5

#### 4.2.1 Isolating set 5

Isolating data set 5 to geocode.

```{r}
unhoused_to_geocode_2 <- unhoused_reformatted |> 
  filter(data_origin == "unhoused_5")

unhoused_to_geocode_2 |> glimpse()
```

#### 4.2.2 Exporting pre-geocode

Exporting data set 5 to geocode in geocodio.

```{r}
unhoused_to_geocode_2 |>
  write_csv("data-processed/unhoused-to-geocode-2.csv")
```

#### 4.2.3 Importing geocoded

Importing geocoded data set 5 from geocodio and cleaning column names.

```{r}
unhoused_geocoded_2 <- read_csv("data-processed/unhoused-geocoded-2.csv") |> 
  clean_names() # the new geocoded columns didn't have clean names

unhoused_geocoded_2 |> glimpse()
```

### 4.4 Merging

Merging the latest batch of geocoded data with the rest of the data set.

#### 4.4.1 Preparing to merge

The data sets have mismatched phone_number data types, this fixes that.

```{r}
unhoused_geocoded_1_fixed <- unhoused_geocoded_1 |> 
  mutate(phone_number = as.character(phone_number))
         
unhoused_geocoded_1_fixed |> glimpse()
```

#### 4.4.2 Merging

```{r}
unhoused_geocoded_merged <- bind_rows(unhoused_geocoded_1_fixed, unhoused_geocoded_2)

unhoused_geocoded_merged |> glimpse()  
```

### 4.5 Removing duplicates

See 4.2.3 for explanation on why this is necessary again.

#### 4.5.1 Finding dupes

This shows both duplicates, so there are actually 130 (260/2) pairs of duplicates.

```{r}
unhoused_geocoded_merged |> 
   get_dupes(!c(data_origin, defendant_address))
```

#### 4.5.2 Removing dupes

Proof it worked: subtract the number of duplicate pairs (130) from the total number of rows (31,537), that is how many rows are outputted below (31,407).

```{r}
unhoused_geocoded_no_dupes <- unhoused_geocoded_merged |>
  arrange(desc(data_origin)) |> # distinct() keeps the first row between two duplicates, so arranging it in descending order by data_origin, it makes sure we keep the most recent row
  distinct(across(!c(
    data_origin, 
    defendant_address)),
    .keep_all = TRUE)

unhoused_geocoded_no_dupes |> glimpse()
```

## 5. OpenRefine

Exporting cleaned data to clean further in OpenRefine. Everything accomplished in OpenRefine is listed in section 5.2.

### 5.1 Exporting for OpenRefine

Exporting clean data to clean more in OpenRefine.

```{r}
unhoused_to_refine <- unhoused_geocoded_no_dupes |> 
  write_csv("data-processed/unhoused-to-refine.csv")

unhoused_to_refine |> glimpse()
```

### 5.2 Importing from OpenRefine {#importing-from-open-refine}

Importing data from OpenRefine. Here are the changes made:

* Clustered misspelled and varied names.
* Created a column that standardizes some judgments more generally.
* Created a column that standardizes some dispositions more generally.
* Created a column indicating whether someone has served jail time.
* Removed all 16 "test" rows.
* Trimmed all leading and trailing whitespace for all columns.
* Made all cell entries uppercase.
* Created a column that standardizes violation descriptions, because starting in October 2025, CC940 and 941 have different violation descriptions than they previously did.

```{r}
unhoused_refined <- read_csv("data-processed/unhoused-refined.csv")
unhoused_refined |> glimpse()
```

## 6. Fixing and grouping dates

### 6.1 Fixing data types

Changing offense_date and judgment_date data types from "dttm" to "date."

```{r}
unhoused_data_types <- unhoused_refined |>
  mutate(offense_date = as_date(offense_date),
         judgment_date = as_date(judgment_date),
        )

unhoused_data_types |> glimpse()
```

### 6.2 Grouping by date

```{r}
unhoused_grouped_dates <- unhoused_data_types |> 
  mutate(off_yr = year(offense_date), # offense year
         off_mo = month(offense_date), # offense month in numbers
         off_mo_label = month(offense_date, label = TRUE), # offense month written
         off_yday = yday(offense_date), # offense day/365
         off_floor_mo = floor_date(offense_date, unit = "month"), # floor month
         off_floor_yr = floor_date(offense_date, "year"), # floor year
         jud_yr = year(judgment_date),
         jud_mo = month(judgment_date),
         jud_mo_label = month(judgment_date, label = TRUE),
         jud_yday = yday(judgment_date),
         jud_floor_mo = floor_date(judgment_date, unit = "month"),
         jud_floor_yr = floor_date(judgment_date, "year")
         )

unhoused_grouped_dates |> glimpse()
```

## 7. Creating and renaming final columns

Creating columns to indicate whether a case is form before/after Whitmire's term began, what type of violation ordinance a violation code is, and a column detailing the amount fined. The final step also renames and selects final columns.

### 7.1 Whitmire flag

Whitmire's term began in January 2024. These columns are separated into offense and judgment because there may be cases where a citation was given before Whitmire's term began, but didn't receive a judgment until after.

#### 7.1.2 Offense date

Creating a column indicating whether an offense occurred before or after Whitmire's term began.

```{r}
unhoused_whitmire_offense <- unhoused_grouped_dates |>
  mutate(
    whitmire_offense = if_else( #if_else function creates a new column based on its argument. it instructs that if the argument is true, use "before," if it's false, use "after." this makes sure we include Jan. 1 2024 in "after"
      offense_date < as.Date("2024-01-01"),
      "BEFORE WHITMIRE",
      "AFTER WHITMIRE"
    )
  )

unhoused_whitmire_offense |> glimpse()
```

#### 7.1.2 Judgment date

Creating a column indicating whether a judgment occurred before or after Whitmire's term began.

```{r}
unhoused_whitmire_judgment <- unhoused_whitmire_offense |>
  mutate(
    whitmire_judgment = if_else(
      judgment_date < as.Date("2024-01-01"),
      "BEFORE WHITMIRE",
      "AFTER WHITMIRE"
    )
  )

unhoused_whitmire_judgment |> glimpse()
```

### 7.2 Violation flag

Making a column to tag whether a citation falls under a civility ordinance vs sidewalk obstruction ordinance.

```{r}
unhoused_citation_tag <- unhoused_whitmire_judgment |> 
  mutate(
    ordinance_type = ifelse(
      violation_code == "CC940" | violation_code == "CC941", "CIVILITY",
      ifelse(
        violation_code == "CC921" | violation_code == "CC942" | violation_code == "CC943" | violation_code == "CC948", "SIDEWALK",
        ifelse(
          violation_code == "CC753" | violation_code == "CC956", "OTHER",
          "OTHER"
  ))))

unhoused_citation_tag |> glimpse()
```

### 7.3 Amount fined

```{r}
unhoused_fined <- unhoused_citation_tag |>
  mutate(amount_fined = total_fine_amount_due +
                        total_fine_amount_paid +
                        total_fine_amount_dismissed)

unhoused_fined |> glimpse()
```

### 7.4 Renaming and selecting columns

Making the final clean object.

```{r}
unhoused_renamed <- unhoused_fined |> 
  select(!c(offense_location)) |> # this column was reworked into full_address, which was used to geocode. offense_location is no longer needed because full_offense_location replaces it
  rename(fines_due = total_fine_amount_due  ,
         fines_paid = total_fine_amount_paid,
         fines_dismissed = total_fine_amount_dismissed,
         full_offense_location = full_address)

unhoused_renamed |> glimpse()
```

## 8. Exporting

Exporting entirely cleaned data for analysis.

```{r}
unhoused_clean <- unhoused_renamed |> 
  write_csv("data-processed/unhoused-clean.csv")
```
