---
title: "Cleaning"
format: html
---

## Goals for this notebook
1. Evaluate data cleanliness
2. Add amount fined column
3. Select columns
4. Remove duplicate rows
5. Clean data using OpenRefine
6. Group by dates
7. Create and select final columns
8. Export clean data

## Setup

Loading the libraries.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(stringr)
```

## Importing

```{r}
unhoused_geocoded <- read_csv("data-processed/unhoused-geocodio-combined.csv")
```

## 1. Evaluating data cleanliness

```{r}
unhoused_geocoded |> glimpse()
```

## 1.1. Cleaning names

```{r}
unhoused_clean_names <- unhoused_geocoded |> 
  clean_names()

unhoused_clean_names |> glimpse()
```

## 1.2. Fixing dates

```{r}
unhoused_dates <- unhoused_clean_names |> #@ i see at least one test row
  mutate(offense_date = as_date(offense_date),
         judgment_date = as_date(judgment_date)
         )

unhoused_dates |> glimpse()
```

## 1.3 Merging misspelled columns

### 1.3.1 Evaluating columns

Looking at columns to evaluate which function to use.

```{r}
unhoused_dates |> 
  select(defendant_address, defendent_address)
```

### 1.3.2 Merging

Merging defendant_address and defendent_address into one column.

```{r}
unhoused_coalesce <- unhoused_dates |>
  mutate(
    defendant_address = coalesce(defendant_address, defendent_address) # this function is able to generate a complete vector from partially missing pieces
  ) |> 
  select(!defendent_address)

unhoused_coalesce |> glimpse()
```

## 2. Adding amount fined column

```{r}
unhoused_fined <- unhoused_coalesce |>
  mutate(amount_fined = total_fine_amount_due +
                          total_fine_amount_paid +
                          total_fine_amount_dismissed)

unhoused_fined |> glimpse()
```

## 3. Selecting columns

```{r}
unhoused_selected <- unhoused_fined |> 
  rename(fines_due = total_fine_amount_due,
         fines_paid = total_fine_amount_paid,
         fines_dismissed = total_fine_amount_dismissed
  ) |> 
  select(!c(phone_number, 
            offense_location)
         )

unhoused_selected |> glimpse()
```

## 4. Remove duplicate rows

### 4.1 Finding dupes

```{r}
unhoused_selected |> 
   get_dupes(data_origin)
```

### 4.2 Removing dupes

```{r}
unhoused_to_refine <- unhoused_selected |>
  arrange(desc(data_origin)) |> # distinct() keeps the first row between two duplicates, so arranging it in descending order by data_origin, it makes sure we keep the most recent row
  distinct(across(!c(
    data_origin, 
    defendant_address)),
    .keep_all = TRUE)

unhoused_to_refine |> glimpse()
```

## 5. OpenRefine

Exporting cleaned data to clean more in OpenRefine. Steps accomplished in OpenRefine listed in the next step.

### 5.1 Exporting for OpenRefine

Exporting clean data to clean more in OpenRefine.

```{r}
unhoused_to_refine |> 
  write_csv("data-processed/unhoused-to-refine.csv")
```

### 5.2 Importing from OpenRefine

Importing data from OpenRefine. Here are the changes made:
* Clustered misspelled names.
* Created a column that standardizes some judgments more generally.
* Created a column that standardizes some dispositions more generally.
* Created a column indicated whether someone has served jail time.
* Removed all 16 "test" rows.

```{r}
unhoused_open_refined <- read_csv("data-processed/unhoused-open-refined.csv")
unhoused_open_refined |> glimpse()
```

## 6. Grouping by dates

```{r}
unhoused_grouped_dates <- unhoused_open_refined |> 
  mutate(
    off_yr = year(offense_date), # offense year
    off_mo = month(offense_date), # offense month in numbers
    off_mo_label = month(offense_date, label = TRUE), # offense month written
    off_yday = yday(offense_date), # offense day/365
    off_floor_mo = floor_date(offense_date, unit = "month"), # floor month
    off_floor_yr = floor_date(offense_date, "year"), # floor year
    jud_yr = year(judgment_date),
    jud_mo = month(judgment_date),
    jud_mo_label = month(judgment_date, label = TRUE),
    jud_yday = yday(judgment_date),
    jud_floor_mo = floor_date(judgment_date, unit = "month"),
    jud_floor_yr = floor_date(judgment_date, "year")
  ) 
  
unhoused_grouped_dates |> glimpse()
```
```{r}
levels(unhoused_grouped_dates$off_mo_label)
```

## 7. Creating and selecting columns

### 7.1 Whitmire flag

I chose to separate these columns because there may be cases where a citation was given before Whitmire's term began, but didn't receive a judgment until after.

#### 7.1.2 Offense date

Creating a column indicating whether an offense occurred before or after Whitmire's term began.

```{r}
unhoused_whitmire_offense <- unhoused_grouped_dates |>
  mutate(
    whitmire_offense = if_else( #if_else function creates a new column based on its argument. it instructs that if the argument is true, use "before," if it's false, use "after." this makes sure we include Jan. 1 2024 in "after"
      offense_date < as.Date("2024-01-01"),
      "BEFORE WHITMIRE",
      "AFTER WHITMIRE"
    )
  )

unhoused_whitmire_offense |> glimpse()
```

#### 7.1.2 Judgment date

Creating a column indicating whether a judgment occurred before or after Whitmire's term began.

```{r}
unhoused_whitmire_judgment <- unhoused_whitmire_offense |>
  mutate(
    whitmire_judgment = if_else(
      judgment_date < as.Date("2024-01-01"),
      "BEFORE WHITMIRE",
      "AFTER WHITMIRE"
    )
  )

unhoused_whitmire_judgment |> glimpse()
```

### 7.2 Violation tag

Making a column to tag whether a citation falls under a civility ordinance vs sidewalk obstruction.

```{r}
unhoused_citation_tag <- unhoused_whitmire_judgment |> 
  mutate(
    citation_type = ifelse(
      violation_code == "CC940" | violation_code == "CC941", "CIVILITY ORDINANCE",
      ifelse(
        violation_code == "CC921" | violation_code == "CC942" | violation_code == "CC943" | violation_code == "CC948", "SIDEWALK OBSTRUCTION ORDINANCE",
        ifelse(
          violation_code == "CC753" | violation_code == "CC956", "OTHER",
          "OTHER"
  ))))

unhoused_citation_tag |> glimpse()
```


### 7.2 Final clean

```{r}
unhoused_clean <- unhoused_citation_tag |> 
  rename(ordinance_type = citation_type)

unhoused_clean |> glimpse()
```

## 8. Exporting

Exporting entirely cleaned data for analysis.

```{r}
unhoused_clean |> 
  write_csv("data-processed/unhoused-clean.csv")
```