---
title: "Analysis"
format: html
code-fold: true
code-summary: "Expand this to see code"
---

## Goals for this notebook

1. Create an object containing only its most recent case number.
2. Investigate citation numbers over time
   i) What is the year with the most citations?
   ii) What is the month with the most citations?
   iii) What is the average number of citations for each month since 2016?
3. Investigate most common violation types
   i) What are the most common violation types of all time?
   ii) What are the most common violation types by year?
   iii) What are the most common violation types by month?
4. Judgement and dispositions
   i) What are the most common judgments?
   ii) What are the most common dispositions?
   iii) How many guilty cases per year?
5. Top people
   i) Who has received the most citations? How many people received multiple citations/fines?
      A.  Who are the most heavily fined people?
      B.  How much of those fines got dismissed?
      C.  How much did they pay?
      D.  How much do they still have due?
      E.  Who has paid the highest amount of dues?
   iii) General fine statistics
      A.  How many people still owe fines to the city?
      B.  How many people have paid their fines?
      C.  How many people have had their fines dismissed?
6. Fines
   i) How many distinct people are there who were fined?
   ii) How many people still owe fines to the city? How much is the city owed in total?
   iii) How many people have paid some/all of their fines?
   iv) How many people have had some/all of their fines dismissed?
7. Jail time
   i) How many jail judgments have been given?
      A.  How many individual people have been jailed?
      B.  How many have been jailed more than once?
   ii) What do monthly "CREDIT TIME SERVED" judgment percentages look like from January 2016 to November 2025?
8. Individual people analysis
   i) Kevin P Shakesnider received the highest jail judgments and fines.
   ii) Patricia Norris has the highest citation count.
9. Mapping analysis
   i) Which zone has had the most citations in this data set?
   ii) How did zone citation counts change by year?
   iii) How did zone citation percentages change by year?

## Set up
```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(sf)
library(mapview)
library(ggspatial)
library(tigris)
library(mapview)
library(plotly)
library(DT)
```

## Importing

```{r}
#| message: false
#| warning: false

unhoused <- read_csv("data-processed/unhoused-clean.csv")
# unhoused |> glimpse()
```

## 1. Creating distinct object {#creating-distinct-object}

Right now, some rows have duplicate case numbers because the second appearance tracks how the case developed. Creating an object that does not have duplicate rows will ensure each row represents an individual citation. The second appearance will be kept because that provides the most updated data.

```{r}
unhoused_distinct <- unhoused |> 
  group_by(case_number) |>  
  arrange(desc(judgment_date)) |> 
  slice(1) |>
  ungroup() |> 
  mutate(off_mo_label = factor(off_mo_label, levels = month.abb, ordered = TRUE), # something changed the mo_label columns from <ord> to <chr>, this fixes that
         jud_mo_label = factor(jud_mo_label, levels = month.abb, ordered = TRUE)
        )

# unhoused_distinct |> glimpse()
```

## 2. Citation numbers over time {#citation-numbers-over-time}

Looking at how individual citation numbers have changed over time.

### 2.1. Year with most citations

What is the year with the most citations?

#### 2.1.1 Making object

```{r}
unhoused_yearly <- unhoused_distinct |> # unhoused_distinct ensures every row counts as an individual citation
  group_by(off_floor_yr, off_yr, whitmire_offense) |> 
  summarize(yearly_citations = n(), .groups = "drop") |> 
  arrange(off_floor_yr) 

unhoused_yearly
```

#### 2.1.2 Plotting

Plotting yearly citations as an interactive bar chart. Note that 2025 is only representative of data until Nov. 20.

```{r}
yearly_plot <- ggplot(
  unhoused_yearly,
  aes(x = off_floor_yr, 
      y = yearly_citations),
  ) +
  geom_col(fill = "#4E79A7") +
  labs(
    title = "Yearly citations fluctuate from 2016 to 2025",
    subtitle = str_wrap("2019 to 2020 showed a 176% increase in citations per year. 2022 had 2007 citations, which showed a roughly 60% decrease from 2020 to 2022."),
    caption = "By Layla Dajani",
    x = "Year",
    y = "Citations per year"
  ) +
  theme_minimal()

yearly_plot |> 
  ggplotly()
```

#### 2.1.3 Data takeaways

* The year with the highest number of citations is 2016 with 5800, followed by 2020 with 5011. The year with the lowest number of citations is 2018, at 1385.	
* 2019 had 1815 citations and 2020 had 5011 -- that is roughly a 176% increase in one year. 2022 had 2007	citations, which is a roughly 60% decrease from 2020 to 2022.

### 2.2 Monthly citations

#### 2.2.1 Jul. 2023 to Nov. 2025

How did the monthly citation count change between Jul. 2023 to Jan. 2025?

##### 2.2.1.1 Making object

```{r}
monthly_citations <- unhoused_distinct |> 
  filter(offense_date >= "2023-07-01") |>  # looking 6 months before Whitmire's term started to see if there is an uptick in citations after Jan. 2024.
  group_by(off_floor_mo, whitmire_offense) |> # adding Whitmire offense for plotting
  summarise(
    monthly_citations = n(),
    .groups = "drop") |> 
    arrange(off_floor_mo)

monthly_citations
```

##### 2.2.1.2 Plotting 

Plotting a bar chart to show the monthly citation count between Jul. 2023 and Nov. 2025. The first dotted line indicates when Whitmire's term began, and the second indicates when he introduced new violation codes. 

```{r}
monthly_citations_plot <- ggplot(monthly_citations, 
       aes(x = off_floor_mo, 
           y = monthly_citations, 
           fill = whitmire_offense)) +
  geom_col() +
  geom_vline(xintercept = ymd("2024-01-01"), 
             linetype = "dashed", 
             color = "grey40") +
  geom_vline(xintercept = ymd("2025-07-01"),
             linetype = "dashed", 
             color = "grey40") +
  scale_fill_manual(
    name = "Before/after Whitmire",
    values = c(
      "BEFORE WHITMIRE" = "#4E79A7", 
      "AFTER WHITMIRE"  = "#E15759"
  )) +
  labs(
    title = "May to October 2025 indicate an upward climb of monthly citation numbers",
    subtitle = str_wrap("From July 2023 to November 2025, October 2025 had the highest citation count at 490 citations. Monthly citations increased by about 463.2% between May and October 2025."),
    x = "Date (yr/mo)",
    y = "Citations per month"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

monthly_citations_plot |> 
  ggplotly()
```

##### 2.2.1.3 Data takeaways

* From July 2023 to November 2025, October 2025 had the highest citation count at 490 citations. This spike is the peak of a monthly increase starting in May 2025. Monthly citations increased by about 463.2% between May and October 2025.
  
#### 2.2.2 Jan. 2016 to Nov. 2025

##### 2.2.2.1 Making object

```{r}
monthly_citations_all <- unhoused_distinct |> 
  group_by(off_floor_mo, whitmire_offense) |> # adding whitmire_offense for plotting
  summarise(
    yr_mo_off = n(),
    .groups = "drop") |>
  arrange(off_floor_mo)

monthly_citations_all
```

##### 2.2.2.2 Plotting

Plotting a line graph to show a monthly citation count between Jan. 2016 and Nov. 2025. The first dotted line indicates when Whitmire's term began, and the second indicates when he introduced new violation codes. 

```{r}
#| warning: false
#| message: false

monthly_citations_all_plot <- ggplot(
  monthly_citations_all, 
  aes(x = off_floor_mo, 
      y = yr_mo_off, 
      color = whitmire_offense)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 0.8) +
  geom_vline(xintercept = ymd("2024-01-01"),
             linetype = "dashed", 
             color = "grey40") +
  geom_vline(xintercept = ymd("2025-07-01"),
             linetype = "dashed", 
             color = "grey40") +
  scale_color_manual(
    values = c(
      "BEFORE WHITMIRE" = "#4E79A7", 
      "AFTER WHITMIRE"  = "#E15759")
    ) +
  labs(
    title = str_wrap("October 2025 was the month with the highest citation count in over four years"),
    subtitle = str_wrap("The first dotted line indicates when Whitmire's term began, and the second indicates when he introduced new violation codes."),
    x = "Year",
    y = "Citations per month",
    color = ""
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("figures/monthly_citations_all_plot.png")

monthly_citations_all_plot |> 
  ggplotly()
```

##### 2.2.2.3 Data takeaways

* From January 2016 to November 2025
  + While October 2025 showed the highest monthly citation count in four years, citation counts peaked even higher in January 2021, at 797 citations, and April 2016, with 715 citations.
  + Monthly citation counts sunk between July 2017 and July 2019. 

### 2.3 Avg monthly citation

What is the average citation count for each calendar month from 2016 to 2025?

#### 2.3.1 Making object

```{r}
mo_avg_citations <- unhoused_distinct |> 
  #filter(off_yr < 2025) |> # i'm including all of 2025, but note that this means Nov. and Dec. are not accurate averages.
  group_by(off_yr,
           off_floor_mo,
           off_mo_label) |> 
  summarise(yr_mo_off = n(),
           .groups = "drop") |> 
  group_by(off_mo_label) |> 
  summarise(avg_mo_off = round(mean(yr_mo_off), 1)) |> 
  arrange(off_mo_label)

mo_avg_citations
```

#### 2.3.2 Plotting

Note that November and December are not accurate averages because 2025 does not contain complete data for those months.

```{r}
mo_avg_citations_plot <- ggplot(
  mo_avg_citations,
  aes(x = off_mo_label, 
      y = avg_mo_off, 
      group = 1)
  ) +
  geom_col(fill = "#4E79A7") +
  labs(
    title = "Monthly citations averaged from Jan. 2016 to Dec. 2025",
    subtitle = str_wrap(""),
    caption = "Graphic by Layla Dajani",
    x = "Month",
    y = "Offenses per month",
  ) +
  theme_minimal()

mo_avg_citations_plot |> 
  ggplotly()
```

#### 2.3.3 Data takeaways

* January is the month that draws the most citations. But surges in specific years could be driving the averages higher in certain months. For instance, January 2016 and 2021 had exceptionally high citation counts, seen in figure 2.2.2.1.

## 3. Violation types {#violation-types}

### 3.1 Most common {#most-common}

Finding the most common violation types.

```{r}
violation_count <- unhoused_distinct |> 
  group_by(violation_description_refined, violation_code) |> 
  summarise(violations = n(),
            .groups = "drop") |> 
  arrange(desc(violations))

violation_count
```

#### 3.1.1 Data takeaways

There are seven types of violations that have been given since 2016. Here is what they mean and their frequency:

1. **CC753 has 29 citations,** a street vendor ordinance
    i) Selling goods in public area: STREET VENDOR (EXPOSE FOR SALE) (SELL) GOODS, TO-WIT:...ON A PUBLIC (SIDEWALK) (STREET) (PROPERTY)
2. **CC921 has 488 citations,** a sidewalk obstruction ordinance
    i) Sidewalk obstruction: OBSTRUCT SIDEWALK BY (PLACING/DEPOSITING/PERMIT PERSON UNDER CONTROL) OBJECT (BOX/MAT./VEH./ETC.)
3. **CC940 has 15,250 citations,** a civility ordinance
    i) Sidewalk obstruction: SIT/LIE DOWN ON (BLANKET/STOOL) PLACED ON SIDEWALK B/W 7:00 AM AND 11:00 PM IN A PROHIBITED AREA
4. **CC941 has 13,723 citations,** a civility ordinance
    i) Sidewalk obstruction: PLACE (BED MAT./PERSONAL POSS.) ON SIDEWALK B/W THE HRS OF 7:00 A.M. & 11:00P.M. IN A PROHIBITED AREA
5. **CC942 has 119 citations,** a sidewalk obstruction ordinance
    i) 	Sidewalk obstruction: IMPAIR OR OBSTRUCT SIDEWALK WITHOUT A PERMIT
6. **CC943 has 9 citations,** a sidewalk obstruction ordinance
    i) Sidewalk obstruction: IMPAIR OR OBSTRUCT SIDEWALK BEYOND SCOPE OF PERMIT
7. **CC948 has 317 citations,** a sidewalk obstruction ordinance
    i) Sidewalk obstruction: PLACE OR ALLOW OBSTRUCTION ON SIDEWALK
8. **CC956 has 1 citation,** a street vendor ordinance
    i) EXPOSE FOR SALE MERCHANDISE, TO-WIT:...ON A PUBLIC (SIDEWALK) (STREET) (ESPLANADE) (PROPERTY)
  
* The most common citation since 2016 is CC940 with 15,250 offenses, with CC941 as a runner-up at 13,723 offenses. The next most common is CC291 with 488 offenses.

### 3.2 Violation codes by year

Plotting yearly violation code totals from Jan. 1, 2016 to Nov. 20, 2025. Note that this means 2025 is an incomplete year.

#### 3.2.1 Making object

```{r}
unhoused_violations <- unhoused_distinct |> 
  group_by(off_floor_yr,
           violation_code)|> 
  summarise(violations = n(),
            .groups = "drop") |> 
  arrange(off_floor_yr)

# unhoused_violations |> glimpse()
```

#### 3.2.2 Plotting

The first dotted line indicates when Whitmire's term began.

```{r}
unhoused_violations_plot <- ggplot(
  unhoused_violations, 
  aes(x = off_floor_yr, 
      y = violations, 
      color = violation_code)) +
  geom_line() +
  geom_point(size = 0.6) +
  geom_vline(xintercept = ymd("2024-01-01"),
             linetype = "dashed", 
             color = "grey70") +
  labs(
    title = str_wrap("Number of citations by year broken down by violation code from Jan. 2016 to Nov. 2025"),
    subtitle = "Violation codes CC753, CC942, CC943, CC948 and CC956 first appeared in 2025.",
    caption = "By Layla Dajani",
    x = "Year",
    y = "Citations per year"
  ) +
  theme_minimal()

unhoused_violations_plot |> 
  ggplotly()
```

#### 3.2.1 Data takeaways

* CC753, CC942, CC943, CC948 and CC956 violation codes first appeared in 2025. 
* CC940 and CC941 seem to follow very similar trends until 2025, when they appear to drastically diverge for the first time in this data set. The missing data from 2025 is not likely impacting this divergence.

### 3.3 Violation codes by month

#### 3.3.1 By count '16 - '25

##### 3.3.1.1 Making object

```{r}
types_by_month_16 <- unhoused_distinct |>
  group_by(off_floor_mo, 
           violation_code,
           ordinance_type) |>
  summarise(citation_count = n(), .groups = "drop") |>
  arrange(off_floor_mo)

types_by_month_16
```

##### 3.3.1.2 Plotting {#plotting-3-3-1-2}

Note that November 2025 is an incomplete month and does not reflect its data wholly. The first dotted line indicates when Whitmire's term began, and the second indicates when he introduced new violation codes.

```{r}
types_by_month_16_plot <- ggplot(
  types_by_month_16,
  aes(x = off_floor_mo, 
      y = citation_count, 
      color = violation_code)) +
  geom_line() +
  geom_point(size = 0.4) +
  geom_vline(xintercept = ymd("2024-01-01"),
             linetype = "dashed", 
             color = "grey70") +
  geom_vline(xintercept = ymd("2025-07-01"),
             linetype = "dashed", 
             color = "grey70") +
  labs(
    title = str_wrap("Number of citations by year broken down by violation code from Jan. 2016 to Nov. 20 2025"),
    subtitle = "",
    caption = "By Layla Dajani",
    x = "Year",
    y = "Citations per year"
      ) +
theme_minimal()

types_by_month_16_plot |> 
  ggplotly()
```

#### 3.3.2 By count '22 - '25 {#3.3.2}

##### 3.3.2.1 Making object

```{r}
types_by_month_22 <- unhoused_distinct |>
  filter(off_floor_mo >= "2022-01-01") |> 
  group_by(off_floor_mo, 
           violation_code,
           ordinance_type) |>
  summarise(citation_count = n(), .groups = "drop") |>
  arrange(off_floor_mo)

types_by_month_22
```

##### 3.3.1.2 Plotting

Note that November 2025 is an incomplete month and does not reflect its data wholly. The first dotted line indicates when Whitmire's term began, and the second indicates when he introduced new violation codes.

```{r}
#| message: false

types_by_month_22_plot <- ggplot(
  types_by_month_22,
  aes(x = off_floor_mo, 
      y = citation_count, 
      color = violation_code)) +
  geom_line() +
  geom_point(size = 0.4) +
  geom_vline(xintercept = ymd("2024-01-01"),
             linetype = "dashed", 
             color = "grey70") +
  geom_vline(xintercept = ymd("2025-07-01"),
             linetype = "dashed", 
             color = "grey70") +
  labs(
    title = str_wrap("Number of citations by year broken down by violation code from Jan. 1, 2022 to Nov. 20, 2025"),
    subtitle = strwrap("The first dotted line indicates when Whitmire's term began, and the second indicates when he introduced new violation codes."),
    caption = "By Layla Dajani",
    x = "Year",
    y = "Citations per year"
      ) +
theme_minimal()

ggsave("figures/types-by-month-22-plot.png")

types_by_month_22_plot |> 
  ggplotly()
```

##### 3.3.1.3 Data takeaways

* This further suggests CC940 and CC941 have followed similar patterns, indicating they have been given out at the same rate. But starting around July 2025, when Whitmire introduced new ordinances, there became less CC941 citations relative to CC940 numbers, while new violation code counts increased. This is further displayed in the following graph.

#### 3.3.3 By percentage

What percentage of CC940 violation codes were met with CC941 violation codes over time? This tracks their correlation.

##### 3.3.3.1 Making object

```{r}
codes_by_pct <- unhoused_distinct |> 
  group_by(off_floor_mo) |> 
  summarise(
    total_citations = n(),
    total_940 = sum(violation_code == "CC940"),
    total_941 = sum(violation_code == "CC941"),
    total_other = sum(violation_code == "CC753", 
                      violation_code == "CC942" | 
                      violation_code == "CC943" | 
                      violation_code == "CC948" | 
                      violation_code == "CC956" | 
                      violation_code == "CC921"),
    cc940_pct = round((total_940/total_citations)*100, 2),
    cc941_pct = round((total_941/total_citations)*100, 2),
    other_pct = round((total_other/total_citations)*100, 2)
  ) |> 
  ungroup() |>
  arrange(off_floor_mo)

codes_by_pct 
```

##### 3.3.3.2 Plotting {#plotting-3-3-3-2}

Note that November 2025 is an incomplete month and does not reflect its data wholly. The first dotted line indicates when Whitmire's term began, and the second indicates when he introduced new violation codes. "ALL OTHERS" includes all codes that are not CC940 or 941.

```{r}
codes_by_pct_plot <- ggplot(
  codes_by_pct,
  aes(x = off_floor_mo)) +
  geom_line(aes(y = cc940_pct,
                color = "CC940")) +
  geom_point(aes(y = cc940_pct,
                color = "CC940"),
             size = 0.4) +
  geom_line(aes(y = cc941_pct,
                color = "CC941")) +
  geom_point(aes(y = cc941_pct,
                color = "CC941"),
             size = 0.4) +
  geom_line(aes(y = other_pct,
                color = "ALL OTHERS")) +
  geom_point(aes(y = other_pct,
                color = "ALL OTHERS"),
             size = 0.4) +
  scale_color_manual(
    name = "VIOLATION CODE",
    values = c(
      "CC941" = "#4E79A7",
      "ALL OTHERS" = "#E15759"
    )
  ) +
  geom_vline(xintercept = ymd("2024-01-01"),
             linetype = "dashed", 
             color = "grey70") +
  geom_vline(xintercept = ymd("2025-07-01"),
             linetype = "dashed", 
             color = "grey70") +
  labs(
    title = str_wrap("CC940, CC941 and all 'other' codes percentage breakdown by month."),
    subtitle = str_wrap(""),
    caption = "By Layla Dajani",
    x = "Year",
    y = "Percentage of total citations"
      ) +
theme_minimal()

codes_by_pct_plot |> 
  ggplotly()
```

##### 3.3.3.3 Data takeaways

* Starting around July 2025, when Whitmire introduced new violation codes, the CC941 percentage decreased, while new violation code percentages increased.
* August 2025 was the first time the percentage of all 'other' codes exceeded CC941 for that month, and October 2025 was they exceeded CC940 for that month.

### 3.4 Ordinance types over time

#### 3.4.1 Making object

Grouping by ordinance type and plotting over time.

```{r}
types_over_time <- unhoused_distinct |>
  filter(offense_date >= "2023-07-01") |> 
  group_by(off_floor_mo, 
           ordinance_type) |>
  summarise(citation_count = n(), .groups = "drop") |>
  arrange(off_floor_mo)

types_over_time
```

#### 3.4.2 Plotting

Note that November 2025 is an incomplete month and does not reflect its data wholly. The first dotted line indicates when Whitmire's term began, and the second indicates when he introduced new violation codes.

```{r}
#| message: false

types_over_time_plot <- ggplot(
  types_over_time,
  aes(x = off_floor_mo, 
      y = citation_count, 
      fill = ordinance_type
      )) +
  geom_col(position = "stack") +
  scale_fill_manual(
    name = "ORDINANCE TYPE",
    values = c(
      "CIVILITY" = "#E15759",
      "SIDEWALK" = "#4E79A7",
      "OTHER" = "#59A14F"
    )
  ) +
  geom_vline(xintercept = ymd("2024-01-01"),
             linetype = "dashed", 
             color = "grey40") +
  geom_vline(xintercept = ymd("2025-07-01"),
             linetype = "dashed", 
             color = "grey40") +
  labs(
    x = "Date",
    y = "Number of citations per month",
    fill = "Violation type",
    title = "Monthly citations by violation type from July 2023 to November 2025",
    subtitle = str_wrap("The first dotted line indicates when Whitmire's term began, and the second indicates when he introduced new violation codes."),
    caption = "Layla Dajani"
      ) +
  theme_minimal()

ggsave("figures/types-over-time-plot.png")

types_over_time_plot |> 
  ggplotly()
```

#### 3.4.3 Data takeaways

October 2025 had highest sidewalk obstruction citations per month in this data set. The 146 citations made up around 30% of the total citations from that month.

## 4. Judgments and dispositions

### 4.1 Most common judgment

#### 4.1.1 Unrefined

What is the most common specific judgment?

```{r}
unhoused |> # not using unhoused_distinct here to show the true judgment stats. even if there are repeated cases, those repeats each received their own judgment in court.
  group_by(judgment) |> 
  summarise(judgment_count = n()) |> 
  arrange(desc(judgment_count))
```

#### 4.1.2 Refined

What is the most common refined judgment?

```{r}
unhoused |>
  group_by(judgment_refined) |> 
  summarise(judgment_count = n()) |>
  arrange(desc(judgment_count))
```

#### 4.1.3 Case dismissed

Break down of case dismissed judgments.

```{r}
unhoused |>
  filter(
    str_detect(judgment, regex("CASE DISMISSED"))
  ) |>
  tabyl(judgment) |>
  rename(judgment_count = n) |>
  mutate(percent = (percent * 100) |> round(1)) |>
  arrange(desc(percent)) |>
  tibble()
```

#### 4.1.4 Data takeaways {#data-takeaways-4-1-4}

* "Case dismissed" is the most common judgment, with a count of 11693.
* The most common reason a case is dismissed is because the officer was not present. Of the 11693 cases that were dismissed, about 72% of cases were dismissed because officers were not present. About 20% of case dismissals were because "OUTREACH(Indigency)."

### 4.2 Most common disposition

What is the most common disposition?

```{r}
unhoused |>
  filter(
    str_detect(judgment, regex("CASE DISMISSED"))
  ) |>
  tabyl(judgment) |>
  rename(judgment_count = n) |>
  mutate(percent = (percent * 100) |> round(1)) |>
  arrange(desc(percent)) |>
  tibble()
```

#### 4.2.1 Data takeaways {#data-takeaways-4-2-1}

* Since "undisposed" means the case still has not been decided, of the cases that have received a disposition, about 59% were dismissed at trial. 
* About 36% of cases that have received a disposition were ruled "guilty by trial judge."

### 4.3 Yearly guilty cases

Guilty disposition counts by year.

#### 4.3.1 Making object

```{r}
yearly_guilty <- unhoused |> 
  filter(disposition == "GUILTY TRIAL BY JUDGE") |> 
  group_by(off_floor_yr) |> 
  summarize(yearly_guilty = n()) |> 
  arrange(off_floor_yr) 

yearly_guilty
```

#### 4.3.2 Plotting

```{r}
yearly_guilty_plot <- ggplot(
  yearly_guilty,
  aes(x = off_floor_yr, 
      y = yearly_guilty
      )) +
  geom_col(fill = "#4E79A7") +
  labs(
    x = "Year",
    y = "Number of guilty judgments per month",
    title = "Guilty judgments by year from 2016 to 2025",
    subtitle = "The dotted line indicates when Whitmire's term began."
      ) +
  theme_minimal()

yearly_guilty_plot |> 
  ggplotly()
```

#### 4.3.1 Data takeaways

* 2018 is the year with the lowest number of guilty dispositions at 50 dispositions.
* 2016 is the year with the highest number of guilty dispositions at 1951 dispositions, followed by 2021 with 931.

## 5. Top people

### 5.1 Distinct people

How many distinct people have received a citation?

```{r}
unhoused_distinct |> 
  summarise(people_count = n_distinct(refined_name))
```

#### 5.1.1 Data takeaway {#data-takeaways-5-1-1}

* There are 4,682 distinct people who have received a citation since 2016. Note that this number is an estimate as the name cleaning process was not perfect.

### 5.2 Most citations

Who has received the most citations since January 2016? 

```{r}
unhoused_distinct |> 
  group_by(refined_name) |> 
  summarise(citation_count = n(),
            first_citation = min(offense_date),
            last_citation = max(offense_date),
            ) |> 
  arrange(desc(citation_count)) |> 
  filter(citation_count >= 100)
```

#### 5.2.1 Data takeaways {#data-takeaways-5-2-1}

* There are 26 people who have received at least 100 total citations. 
* The person with the most citations is Patricia Norris, who has received 794 total citations, all given between October 2017 and May 2025. 
* The person with the second most citations is Forest H Smith, who has received 351 total citations, all given between August 2020 and November 2025.

### 5.3 Most fines and dues {#most-fines-and-dues}

* Who are the most heavily fined people? 
* How much of those fines got dismissed? 
* How much did they pay? 
* How much do they still have due? Who has paid the highest amount of dues? 

fines_paid_not_credit is how much the person paid when time_served = "NO."
fines_paid_served_credit is how much the person paid when time_served = "YES."

```{r}
individual_fines <- unhoused_distinct |>
  group_by(refined_name) |>
  summarise(
    citations = n(),
    total_fined = sum(amount_fined, na.rm = TRUE) |> round(0),
    total_fines_dismissed = sum(fines_dismissed, na.rm = TRUE) |> round(0),
    total_fines_paid = sum(fines_paid, na.rm = TRUE) |> round(0),
    fines_paid_not_credit = sum(fines_paid[time_served == "NO"], na.rm = TRUE) |> round(0),
    fines_paid_served_credit = sum(fines_paid[time_served == "YES"], na.rm = TRUE) |> round(0),
    total_fines_due = sum(fines_due, na.rm = TRUE) |> round(0)
  ) |>
  arrange(desc(total_fined))

individual_fines |> 
  datatable()
```

#### 5.3.1 Data takeaways {#data-takeaways-5-3-1}

* Most fined:
  + Patricia Norris has been charged the highest total fines at around \$199,123. She hasn't paid any of her fines, but has had about \$149,118 dismissed, which leaves about \$50,006 that she still owes. 
  + Forest H Smith is the second most fined person, with around \$86,223 total fines. He has also never paid any of his fines, but has had roughly \$9,584 dismissed, leaving around \$76,640 that he still owes.
* Paid the most in cash:
  + Kevin P Shakesnider is the person who has paid the highest amount of fines when considering cash and time served combined. He is also the person who has paid the most fines in cash. He has paid about \$26,108 of his total fined \$45,125 with "CREDIT TIME SERVED," he paid \$2576 in cash, and has had roughly \$10,916 dismissed. As of November 2025, he still owes about \$5,525.
  + 14 people have paid over $1,000 in cash. The top three in descending order are Shakesnider, Melissa Clark and Robin Harrison.
* Paid the most in time served:
  + Bobby L Anderson Jr. has paid the most fines with "CREDIT TIME SERVED." He has been fined roughly \$38,245. Of that, he paid \$26,489 with time served, none in cash and had about \$9,287 dismissed.

## 6. Fines

1. How many distinct people are there who were fined?
2. How many people still owe fines to the city? How much is the city owed in total?
3. How many people have paid some/all of their fines when time_served = "NO"?
4. How many people have paid some/all of their fines when time_served = "YES"?
5. How many people have had some/all of their fines dismissed?

### 6.1 Fined people

How many distinct people are there who were fined?

```{r}
individual_fines |> 
  filter(total_fined > 0) |>
  summarise(total_fined_people = n(),
            amount_fined_total = sum(total_fined),
            percent_of_total_people = round((total_fined_people / 4598) * 100, 1)
            )
```

### 6.2 Owed fines

How many people still owe fines to the city? How much is the city owed in total?

```{r}
individual_fines |> 
  filter(total_fines_due > 0) |> 
  summarise(people_owing_fines = n(),
            owed_fines_total = sum(total_fines_due),
            percent_of_people = round((people_owing_fines / 4456) * 100, 1),
            percent_of_fines = round((owed_fines_total / 6735088) * 100, 1)
  )
```

### 6.3 Fines paid not served

How many people have paid some/all of their fines when time_served = "NO"?

#### 6.3.1 Some fines paid

How many people have paid at least some of their fines when time_served = "NO"?

```{r}
individual_fines |> 
  filter(fines_paid_not_credit > 0) |> 
  summarise(people_paid_some_fines = n(),
            paid_fines_not_credit_total = sum(fines_paid_not_credit),
            percent_of_people = round((people_paid_some_fines / 4456) * 100, 1),
            percent_of_fines = round((paid_fines_not_credit_total / 6735088) * 100, 1)
  )
```

#### 6.3.2 All fines paid - no dismissals

How many people have paid all of their fines when time_served = "NO" with no help from dismissals?

```{r}
individual_fines |> 
  filter(total_fined > 0,
         fines_paid_not_credit == total_fined
  ) |> 
  summarise(people_paid_all_fines = n(),
            paid_fines_not_crdit_total = sum(fines_paid_not_credit),
            percent_of_people = round((people_paid_all_fines / 4456) * 100, 1),
            percent_of_fines = round((paid_fines_not_crdit_total / 6735088) * 100, 1)
  )
```

### 6.4 Fines paid time served

How many people have paid some/all of their fines when time_served = "YES"?

#### 6.4.1 Some fines paid

How many people have paid at least some of their fines when time_served = "YES"?

```{r}
individual_fines |> 
  filter(fines_paid_served_credit > 0) |> 
  summarise(people_paid_some_fines = n(), 
            paid_fines_credit_total = sum(fines_paid_served_credit),
            percent_of_people = round((people_paid_some_fines / 4456) * 100, 1),
            percent_of_fines = round((paid_fines_credit_total / 6735088) * 100, 1)
  )
```

#### 6.4.2 All fines paid - no dismissals

How many people have paid all of their fines when time_served = "YES" with no help from dismissals?

```{r}
individual_fines |> 
  filter(total_fined > 0,
         fines_paid_served_credit == total_fined
  ) |> 
  summarise(people_paid_all_fines = n(),
            paid_fines_credit_total = sum(fines_paid_served_credit),
            percent_of_people = round((people_paid_all_fines / 4456) * 100, 1),
            percent_of_fines = round((paid_fines_credit_total / 6735088) * 100, 1)
  )
```

### 6.5 Fines dismissed

How many people have had some/all of their fines dismissed?

#### 6.5.1 Some dismissed

How many people have had at least some of their fines dismissed?

```{r}
individual_fines |> 
  filter(total_fines_dismissed > 0) |> 
  summarise(people_some_dismissed_fines = n(),
            total_dismissed = sum(total_fines_dismissed),
            percent_of_people = round((people_some_dismissed_fines / 4456) * 100, 1),
            percent_of_fines = round((total_dismissed / 6735088) * 100, 1)
  )
```

#### 6.5.2 All dismissed

How many people have had all of their fines dismissed?

```{r}
individual_fines |> 
  filter(total_fines_dismissed > 0,
         total_fines_dismissed == total_fined
  ) |> 
  summarise(people_all_dismissed_fines = n(),
            total_dismissed = sum(total_fines_dismissed),
            percent_of_people = round((people_all_dismissed_fines / 4456) * 100, 1),
            percent_of_fines = round((total_dismissed / 6735088) * 100, 1)
  )

```

### 6.6 Data takeaways {#data-takeaways-6-6}

1. **Total fined:** There are about 4,456 out of the total 4,598 distinct people who have received a fine with their citation since 2016, their cumulative total being $6,735,614.
    i) So about 97% who have received citations have also received a fine at some point.
2. **Total owed:** Of the 4,456 people, 2,972 still owe fines. The sum of total fines still owed to the city is \$3,013,654.
    i) So the city is still owed about 45% of the roughly $6.7 million the city has fined unhoused people since 2016.
3. **Paid time_served "YES":** Of the 4,456 people, 227 have paid at least some of their fines when time_served = "YES." 78 of those people have paid off all their fines without any help from dismissals. The total amount of fines paid when time_served = "YES" since 2016 is about \$80,039.
    i) So just over 1% of the roughly $6.7 million the city has fined unhoused people since 2016 has been paid in cash.
4. **Paid time_served "NO":** Of the 4,456 people, 1066 have paid at least some of their fines when time_served = "NO." 467 of those people have paid off all their fines without any help from dismissals. The total amount of fines paid when time_served = "NO" since 2016 is about \$1,406,782.
    i) So about 21% of the roughly $6.7 million the city has fined unhoused people since 2016 has been paid in time served.
5. **Total dismissed:** Of the 4,456 people, 1459 have had at least some of their fines dismissed, and 699 of those 1,459 people had all of their fines dismissed. A total of \$2,235,140	has been dismissed. 
    i) So about 33% of the roughly $6.7 million the city has fined unhoused people since 2016 has been dismissed.

## 7. Jail time

### 7.1 Credit time served

#### 7.1.1 Distinct jailed judgments

How many individual people have received at least one CREDIT TIME SERVED judgment? Who has received the most CREDIT TIME SERVED judgments?

```{r}
unhoused |> 
  group_by(refined_name) |> 
  summarise(credit_jail_judgments = sum(time_served == "YES")) |> 
  filter(credit_jail_judgments > 1) |> 
  arrange(desc(credit_jail_judgments))
```

#### 7.1.2 Distinct judgment dates

Who has received the most CREDIT TIME SERVED judgments on distinct judgment dates? 

**Explanation:** oftentimes, courts rule on several citations for the same person on one day. For example, Kevin Shakesnider didn't receive all 136 of his CREDIT TIME SERVED judgments on 136 separate days, he was assigned the same judgment date for several of his citations.

```{r}
unhoused |> 
  filter(time_served == "YES") |> 
  distinct(refined_name, judgment_date) |> 
  group_by(refined_name) |> 
  summarise(credit_jail_judgment_days = n()) |> 
  #filter(credit_jail_judgment_days > 1) |> 
  arrange(desc(credit_jail_judgment_days))
```

#### 7.1.3 Data takeaways

* Of the 4,598 people who have received a citation since 2016, 1,066 of them received a "CREDIT TIME SERVED" judgment at least once. Of those 4,598, 845 people have received more than one "CREDIT TIME SERVED" judgment.
* The person with the highest "CREDIT TIME SERVED" judgment count is Kevin P Shakesnider, who has received 136  judgments. Those 136 judgments were given across 22 distinct judgment dates.
* The person with the second highest "CREDIT TIME SERVED" judgment count is BOBBY L ANDERSON JR, who has received 104  judgments. Those 104 judgments were given across 7 distinct judgment dates.
* Note that although Patricia Norris has received the highest citation count, she has never received a "CREDIT TIME SERVED" judgment.

## 8. Individual people analysis

### 8.1. Kevin P Shakesnider

Shakesnider data takeaways so far:

* **Highest paid cash and "cash + time served" fines:** Kevin P Shakesnider is the person who has paid the highest amount of fines when considering cash and time served combined. He is also the person who has paid the most fines in cash. He has been fined a total of \$45,125. He paid about \$26,108 with "CREDIT TIME SERVED," he paid \$2576 in cash, and has had roughly \$10,916 dismissed. As of November 2025, he still owes about \$5,525.
* **Highest "CREDIT TIME SERVED" judgments:** He is also the person with the highest "CREDIT TIME SERVED" judgments at 136. Those 136 judgments were given across 22 distinct judgment dates.

#### 8.1.1 "CREDIT TIME SERVED"

When did Kevin P Shakesnider receive the highest jail citations in one judgment? What is the overall date range?

```{r}
shakesnider_jailed <- unhoused |> 
  filter(refined_name == "SHAKESNIDER, KEVIN P") |>
  group_by(judgment_date) |>
  summarise(jail_count = n()) |> 
  arrange(desc(jail_count))

shakesnider_jailed
```

#### 8.1.2 Data takeaway

All 21 of his judgment dates fall between Jan. 13 2016 and Nov. 25 2015. On June 23, he received a "CREDIT TIME SERVED" judgment for 26 citations, his highest in one single day.

### 8.2 Patricia Norris {#patricia-norris}

Norris takeaway so far:

* Patricia Norris has been charged the highest total fines at around \$199,123. She hasn't paid any of her fines, but has had about \$149,118 dismissed, which leaves about \$50,006 that she still owes. 

#### 8.2.1 Visualizing

Making Norris object.

```{r}
norris <- unhoused |> 
  filter(refined_name == "NORRIS, PATRICIA")

norris
```

Checking for what kinds of violation codes Norris has been given.

```{r}
norris |> 
  distinct(violation_code)
```

Making the object to plot.

```{r}
norris_citations <- norris |> 
  group_by(offense_date) |> 
  summarise(citation_count = n()) |>
  arrange(offense_date)

norris_citations
```

##### 8.2.2.2 Plotting

```{r}
#| message: false

norris_citations_plot <- ggplot(
  norris_citations, 
       aes(x = offense_date, 
           y = citation_count)) +
  geom_point(alpha = 1 / 3) +
  labs(
    title = "Patricia Norris citation count by offense date",
    x = "Offense date",
    y = "Citation count",
    caption = "By Layla Dajani"
  ) +
  theme_minimal()

ggsave("figures/norris_citations_plot.png")

norris_citations_plot |> ggplotly()
```

###### 8.2.2.1.1 Data takeaways

* Each one of these points is a day, so the higher the citation count, the more citations she received in a single day. Almost every time Patricia Norris has been given a citation, she has also been given at least one other citation. She was also hit with citations almost consistently from January 2020 to May 2025.
* Norris has only ever received violation codes CC940 and CC941.

## 9. Mapping analysis

Mapping citations by ordinance zones. Skip to section 9.2 for results.

### 9.1 Preparing to plot map

Setting up to map.

#### 9.1.1 Importing district data

Importing Houston district data as an sf object.

```{r}
#| message: false
#| warning: false

hou_zones <- st_read(
  "additional-materials/civility-ordinance-boundary.geojson"
)

# hou_zones
```

#### 9.1.2 Simple feature

Turning the addresses in a Simple Feature Collection.

```{r}
pts_zones <- st_as_sf(
  unhoused_distinct,
  coords = c("geocodio_longitude", "geocodio_latitude"),
  crs = 4326
)

# pts_zones |> glimpse()
```

#### 9.1.3 Fixing invalid district lines

```{r}
clean_zones <- st_make_valid(hou_zones)

cleaner_zones <- st_transform(clean_zones, 4326)
```

#### 9.1.4 Citation to zone

Assigning each citation to a zone.

```{r}
#| message: false

zoning_citations <- st_join(pts_zones, cleaner_zones, join = st_within)

# zoning_citations
```

Cleaning names after joining:

```{r}
zoning_citations_clean <- zoning_citations |> 
  clean_names()

# zoning_citations_clean |> glimpse()
```

#### 9.1.5 Citations by zone

Which zone has received the most citations from 2016 to Sept. 2025? 

Grouping data by zone, preparing to map.

```{r}
zone_counts <- zoning_citations_clean |>
  st_drop_geometry() |>
  count(name, name = "citation_count")

zone_counts
```


### 9.2 Plotting {#plotting}

Cleaning cleaner_zone names and joining.

```{r}
cleaner_names_zones <- clean_names(cleaner_zones)
 
zones_to_map <- cleaner_names_zones |>
  left_join(zone_counts, by = "name")
```

#### 9.2.1 General mapview

Plotting an interactive heat map to show each district's citations.

```{r}
mapview(
  zones_to_map,
  zcol = "citation_count",
  popup = c("name", "citation_count"),
  layer.name = "Citations"
) 
```

#### 9.2.2 Zones over time

Visualizing each zone's citation count by year in stacked bar charts, one by citation count and the other by citation percentage.

##### 9.2.2.1 By count

###### 9.2.2.1.1 Making object

```{r}
zone_year_counts <- zoning_citations_clean |>
  st_drop_geometry() |>
  group_by(off_yr, name) |>
  summarise(citation_count = n(),
            .groups = "drop") |> 
  arrange(desc(citation_count))

zone_year_counts
```

###### 9.2.2.1.2 Plotting

```{r}
zone_year_counts_plot <- ggplot(
  zone_year_counts, 
  aes(x = factor(off_yr), 
      y = citation_count, 
      fill = name)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(
  "#E41A1C",
  "#377EB8",
  "#4DAF4A",
  "#FF7F00",
  "#984EA3",
  "#FFFF33",
  "#A65628",
  "#F781BF",
  "#999999",
  "#66C2A5",
  "#FC8D62"   
  )) +
  labs(
    x = "Year",
    y = "Citation count",
    fill = "Zone",
    title = "Citation counts per zone by year",
    caption = "By Layla Dajani"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

zone_year_counts_plot |> 
  ggplotly()
```

##### 9.2.2.2 By percentage

###### 9.2.2.2.1 Making object

```{r}
zone_year_pct <- zone_year_counts |>
  group_by(off_yr) |>
  mutate(citation_pct = round(citation_count / sum(citation_count) * 100, 1)) |> 
  arrange(desc(citation_pct))

zone_year_pct
```

###### 9.2.2.2.2 Plotting

```{r}
zone_year_pct_plot <- ggplot(
  zone_year_pct, 
  aes(x = factor(off_yr), 
      y = citation_pct, 
      fill = name)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(
  "#E41A1C",
  "#377EB8",
  "#4DAF4A",
  "#FF7F00",
  "#984EA3",
  "#FFFF33",
  "#A65628",
  "#F781BF",
  "#999999",
  "#66C2A5",
  "#FC8D62"   
  )) +
  labs(
    x = "Year",
    y = "Percentage",
    fill = "Zone",
    title = "Citation percentages per zone by year",
    caption = "By Layla Dajani"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

zone_year_pct_plot |> 
  ggplotly()
```
