---
title: "Analysis"
format: html
code-fold: true
code-summary: "Expand this to see code"
---

## Goals for this notebook
1. Create an object containing only rows with only its most recent case number.
2. Investigate citation numbers over time
    i) What is the year with the most citations?
    ii) What is the month with the most citations?
    iii) What is the average number of citations for each month since 2016?
3. Investigate most common violation types
    i) What are the most common violation types of all time?
    ii) What are the most common violation types by year?
4. Judgement and dispositions
    i) What is the most common judgement?
    ii) What is the most common disposition?
    iii) How many guilty cases per year?
    iv) What is the month with the most guilty dispositions and highest percentage?
    v) What is the average percentage of guilty dispositions for each month from 2016-2024?
5. Top people
    i) Who has received the most citations? How many people received multiple citations/fines?
    ii) Individual fine statistics
        A. Who are the most heavily fined people?
        B. How much of those fines got dismissed? 
        C. How much did they pay? 
        D. How much do they still have due? 
        E. Who has paid the highest amount of dues?
    iii) General fine statistics
        A. How many people still owe fines to the city?
        B. How many people have paid their fines?
        C. How many people have had their fines dismissed?
6. Fines
    i) How many distinct people are there who were fined?
    ii) How many people still owe fines to the city? How much is the city owed in total?
    iii) How many people have paid some/all of their fines?
    iv) How many people have had some/all of their fines dismissed?
7. Jail time
    i) How many jail judgments have been given?
    ii) How many individual people have been jailed? How many have been jailed more than once?
8. Individual people analysis
    i) Kevin Shakesnider received the highest jail judgments and fines, let's visualize that.
    ii) Patricia Norris has the highest citation count, let's visualize that.
9. Mapping analysis

## Set up
```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(sf)
library(mapview)
library(ggspatial)
library(tigris)
library(mapview)
library(plotly)
```

## Importing

```{r}
unhoused <- read_csv("data-processed/unhoused-clean.csv")
unhoused |> glimpse()
```

## 1. Creating distinct object

Right now, some rows have duplicate case numbers because the second appearance tracks how the case developed. Creating an object that does not have duplicate rows will ensure each row represents an individual citation. The second appearance will be kept because that provides the most updated data.

```{r}
unhoused_distinct <- unhoused |> 
  group_by(case_number) |>  
  arrange(desc(judgment_date)) |> 
  slice(1) |>
  ungroup() |> 
  mutate(off_mo_label = factor(off_mo_label, levels = month.abb, ordered = TRUE), # something changed the mo_label columns from <ord> to <chr>, this fixes that
         jud_mo_label = factor(jud_mo_label, levels = month.abb, ordered = TRUE)
        )

unhoused_distinct |> glimpse()
```

## 2. Citation numbers over time

Looking at how individual citation numbers have changed over time.

### 2.1. Year with most citations

What is the year with the most citations?

#### 2.1.1 Making object

```{r}
unhoused_yearly <- unhoused_distinct |> # unhoused_distinct ensures every row counts as an individual citation
  group_by(off_floor_yr, off_yr, whitmire_offense) |> 
  summarize(yearly_offenses = n()) |> 
  arrange(off_floor_yr) 

unhoused_yearly
```

#### 2.1.2 Plotting

Plotting yearly offenses as an interactive bar chart.

```{r}
yearly_plot <- ggplot(
  unhoused_yearly,
  aes(x = off_floor_yr, 
      y = yearly_offenses),
  ) +
  geom_col(fill = "#4E79A7") +
  labs(
    title = "Yearly citations fluctuate from 2016 to 2025",
    subtitle = "Note that 2025 is only representative of data until Nov. 20.",
    caption = "By Layla Dajani",
    x = "Year",
    y = "Offenses per year"
  ) +
  theme_minimal()

ggsave("figures/yearly-plot.png")

yearly_plot |> 
  ggplotly()
```

#### 2.1.3 Data takeaways

* The year with the highest number of citations is 2016 with 5800, followed by 2020 with 5011. The year with the lowest number of citations is 2018, at 1385.	
* 2019 saw 1815 citations and 2020 had 5011 -- that is roughly a 176% increase in one year. 2022 had 2007	citations, which is a roughly 60% decrease from 2020 to 2022.

### 2.2 Highest citation month

#### 2.2.1 Jul. 2023 to Nov. 2025

What's the month with the most citations from Jul. 2023 to Jan. 2025?

```{r}
monthly_citations <- unhoused_distinct |> 
  filter(offense_date >= "2023-07-01") |>  # looking 6 months before Whitmire's term started to see if there is an uptick in citations after Jan. 2024.
  group_by(off_floor_mo, whitmire_offense) |> # adding Whitmire offense for plotting
  summarise(
    yr_mo_off = n()) |> 
    arrange(off_floor_mo)

monthly_citations
```

##### 2.2.1.1 Plotting - for summary

@ Include in summary.

Plotting a bar chart to show a monthly citation count between Jul. 2023 and Nov. 2025. Note that the Nov. 2025 data only goes to Nov. 20, and therefore cannot be accurately compared to all other months.  

```{r}
monthly_citations_plot <- ggplot(monthly_citations, 
       aes(x = off_floor_mo, 
           y = yr_mo_off, 
           fill = whitmire_offense)) +
  geom_col() +
  geom_vline(xintercept = ymd("2024-01-01"), 
             linetype = "dashed", 
             color = "grey40") +
  geom_vline(xintercept = ymd("2025-07-01"),
             linetype = "dashed", 
             color = "grey40") +
  scale_fill_manual(
    name = "Before/after Whitmire",
    values = c(
      "BEFORE WHITMIRE" = "#4E79A7", 
      "AFTER WHITMIRE"  = "#E15759"
  )) +
  labs(
    title = "May to October 2025 indicate an upward climb of monthly citation numbers",
    subtitle = str_wrap("The first dotted line indicates when Whitmire's term began, and the second indicates when he introduced new violation codes. Note that November 2025 only includes data through Nov. 20."),
    x = "Date (yr/mo)",
    y = "Citations per month"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("figures/monthly_citations_plot.png")

monthly_citations_plot |> 
  ggplotly()
```

##### 2.2.1.2 Data takeaways

* Since July 2023, 6 months before Whitmire's term began: 
  + The month with the highest citation number since July 2023 is October 2025 with 490 citations, followed by September 2025 with 338 citations. 
  + The month with the lowest citation number since July 2023 is July 2023, with 66 citations. This is followed by May 2023 with 87 citations.
  
#### 2.2.2 Jan. 2016 to Nov. 2025

```{r}
monthly_citations_all <- unhoused_distinct |> 
  group_by(off_floor_mo, whitmire_offense) |> # adding whitmire_offense for when i plot
  summarise(
    yr_mo_off = n()) |>
  arrange(off_floor_mo)

monthly_citations_all
```

##### 2.2.2.1 Plotting - for summary

Plotting a line graph to show a monthly citation count between Jan. 2016 and Nov. 2025. Note that the Nov. 2025 data only goes to Nov. 20, and therefore cannot be accurately compared to all other months. 

```{r}
monthly_citations_all_plot <- ggplot(monthly_citations_all, 
  aes(x = off_floor_mo, 
      y = yr_mo_off, 
      color = whitmire_offense)) +
  geom_line(size = 0.9) +
  geom_point(size = 0.8) +
  geom_vline(xintercept = ymd("2024-01-01"),
             linetype = "dashed", 
             color = "grey40") +
  geom_vline(xintercept = ymd("2025-07-01"),
             linetype = "dashed", 
             color = "grey40") +
  scale_color_manual(
    values = c(
      "BEFORE WHITMIRE" = "#4E79A7", 
      "AFTER WHITMIRE"  = "#E15759")
    ) +
  labs(
    title = "Oct. 2025 was the month with the highest citation count in over four years",
    subtitle = str_wrap("Note that November 2025 only includes data through Nov. 20."),
    x = "Year",
    y = "Citations per month",
    color = ""
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("figures/monthly_citations_all_plot.png")

monthly_citations_all_plot |> 
  ggplotly()
```

##### 2.2.2.2 Data takeaways

* Since 2016
  + The month with the highest citation number since 2016 is January 2021, with 797 citations. This is followed by April 2016 with 715 citations.
  + The month with the lowest citation number since 2016 is March 2022 with 29 citations. This is followed by December 2018 with 37 citations.

### 2.3 Avg monthly citation

#### 2.3.1 Creating object

What is the average monthly citation count for each month from 2016 to 2024? Note that this means November and all of December 2025 are not accurate averages.

```{r}
mo_avg_citations <- unhoused_distinct |> 
  #filter(off_yr < 2025) |> #@ i'm including all of 2025, but note that this means Nov. and Dec. are not accurate averages.
  group_by(off_yr,
           off_floor_mo,
           off_mo_label) |> 
  summarise(yr_mo_off = n()) |> 
  group_by(off_mo_label) |> 
  summarise(avg_mo_off = round(mean(yr_mo_off), 1)) |> 
  arrange(off_mo_label)

mo_avg_citations
```

#### 2.3.2 Plotting

```{r}
mo_avg_citations_plot <- ggplot(
  mo_avg_citations,
  aes(x = off_mo_label, 
      y = avg_mo_off, 
      group = 1)
  ) +
  geom_col(fill = "#4E79A7") +
  labs(
    title = "Monthly citations averaged from Jan. 2016 to Dec. 2024",
    subtitle = str_wrap(""),
    caption = "Graphic by Layla Dajani",
    x = "Month",
    y = "Offenses per month",
  ) +
  theme_minimal()

mo_avg_citations_plot |> 
  ggplotly()
```

#### 2.3.3 Data takeaways

* May and June seem to be the months that draw the most citatons. Surges in specific years could be driving the averages higher in certain months. For instance, January 2021 had the highest citations (787) of any month since 2016.

## 3. Violation types

### 3.1 Most common

Finding the most common violation types.

```{r}
violation_count <- unhoused_distinct |> 
  group_by(violation_description_refined, violation_code) |> 
  summarise(violations = n()) |> 
  arrange(desc(violations))

violation_count
```

#### 3.1.1 Data takeaways

There are seven types of violations that have been given since 2016. Here is what they mean and their frequency:

1. **CC753 has 29 citations,** a street vendor ordinance
    i) Selling goods in public area: STREET VENDOR (EXPOSE FOR SALE) (SELL) GOODS, TO-WIT:...ON A PUBLIC (SIDEWALK) (STREET) (PROPERTY)
2. **CC921 has 488 citations,** a sidewalk obstruction ordinance
    i) Sidewalk obstruction: OBSTRUCT SIDEWALK BY (PLACING/DEPOSITING/PERMIT PERSON UNDER CONTROL) OBJECT (BOX/MAT./VEH./ETC.)
3. **CC940 has 15,250 citations,** a civility ordinance
    i) Sidewalk obstruction: SIT/LIE DOWN ON (BLANKET/STOOL) PLACED ON SIDEWALK B/W 7:00 AM AND 11:00 PM IN A PROHIBITED AREA
4. **CC941 has 13,723 citations,** a civility ordinance
    i) Sidewalk obstruction: PLACE (BED MAT./PERSONAL POSS.) ON SIDEWALK B/W THE HRS OF 7:00 A.M. & 11:00P.M. IN A PROHIBITED AREA
5. **CC942 has 119 citations,** a sidewalk obstruction ordinance
    i) 	Sidewalk obstruction: IMPAIR OR OBSTRUCT SIDEWALK WITHOUT A PERMIT
6. **CC943 has 9 citations,** a sidewalk obstruction ordinance
    i) Sidewalk obstruction: IMPAIR OR OBSTRUCT SIDEWALK BEYOND SCOPE OF PERMIT
7. **CC948 has 317 citations,** a sidewalk obstruction ordinance
    i) Sidewalk obstruction: PLACE OR ALLOW OBSTRUCTION ON SIDEWALK
8. **CC956 has 1 citation,** a street vendor ordinance
    i) EXPOSE FOR SALE MERCHANDISE, TO-WIT:...ON A PUBLIC (SIDEWALK) (STREET) (ESPLANADE) (PROPERTY)
  
* The most common citation since 2016 is CC940 with 15,250 offenses, with CC941 as a runner-up at 13,723 offenses. The next most common is CC291 with 488 offenses.

### 3.2 Violation codes by year

Plotting yearly violation code totals from Jan. 1, 2016 to Nov. 20, 2025. Note that this means 2025 is an incomplete year.

#### 3.2.1 Create object

```{r}
unhoused_violations <- unhoused_distinct |> 
  group_by(off_floor_yr,
           violation_code)|> 
  summarise(violations = n()) |> 
  arrange(off_floor_yr)

unhoused_violations |> glimpse()
```

### 3.2.2 Plotting -- for summary

@for summary

Note that 2025 is an incomplete year, meaning is not representative of the whole year and cannot be reliably compared to past complete years.

```{r}
unhoused_violations_plot <- ggplot(
  unhoused_violations, 
  aes(x = off_floor_yr, 
      y = violations, 
      color = violation_code)) +
  geom_line() +
  geom_point(size = 0.6) +
  geom_vline(xintercept = ymd("2024-01-01"),
             linetype = "dashed", 
             color = "grey70") +
  geom_vline(xintercept = ymd("2025-07-01"),
             linetype = "dashed", 
             color = "grey70") +
  labs(
    title = str_wrap("Number of citations by year broken down by violation code from Jan. 2016 to Nov. 2025"),
    subtitle = "",
    caption = "By Layla Dajani",
    fill = "Citations",
    x = "Year",
    y = "Citations per year"
  ) +
  theme_minimal()

ggsave("figures/unhoused_violations_plot.png")

unhoused_violations_plot |> 
  ggplotly()
```

#### 3.2.1 Data takeaways

* CC753, CC942, CC943, CC948 and CC956 violation codes first appeared in 2025. 
* CC940 and CC941 seem to follow very similar trends until 2025, when they appear to drastically diverge for the first time in this data set. The missing data from 2025 is not likely impacting this divergence.

### 3.3 Violation codes by month

#### 3.3.1 By number

##### 3.3.1.1 Making object

```{r}
types_by_month <- unhoused_distinct |>
  #filter(offense_date <= "2025-10-31") |> # note that this not filtering means Nov. 2025 is included, despite being an incomplete month
  group_by(off_floor_mo, 
           violation_code,
           ordinance_type) |>
  summarise(citation_count = n(), .groups = "drop") |>
  arrange(off_floor_mo)

types_by_month
```

##### 3.3.1.2 Plotting

```{r}
types_by_month_plot <- ggplot(
  types_by_month,
  aes(x = off_floor_mo, 
      y = citation_count, 
      color = violation_code)) +
  geom_line() +
  geom_point(size = 0.4) +
  geom_vline(xintercept = ymd("2024-01-01"),
             linetype = "dashed", 
             color = "grey70") +
  geom_vline(xintercept = ymd("2025-07-01"),
             linetype = "dashed", 
             color = "grey70") +
  labs(
    title = str_wrap("Number of citations by year broken down by violation code from Jan. 2016 to Nov. 20 2025"),
    subtitle = "",
    caption = "By Layla Dajani",
    x = "Year",
    y = "Citations per year"
      ) +
theme_minimal()

types_by_month_plot |> 
  ggplotly() #@ add new column with descriptors 1 word
```

##### 3.3.1.3 Data takeaways

* This further suggests CC940 and CC941 lines follow similar patterns, indicating they have been given out at the same rate. But starting around July 2025, when Whitmire introduced new ordinances, there became less CC941 citations relative to CC940 numbers, while new ordinance citations increased. This is further displayed in the following graph.

#### 3.3.2 By CC941 percentage

What percentage of CC940 violation codes were met with CC941 violation codes over time? This tracks their correlation.

##### 3.3.2.1 Making object

```{r}
codes_by_pct <- unhoused_distinct |> 
  #filter(violation_code == "CC940" | violation_code == "CC941") |> 
  group_by(off_floor_mo) |> 
  summarise(
    total_940 = sum(violation_code == "CC940"),
    total_941 = sum(violation_code == "CC941"),
    total_other = sum(violation_code == "CC753", 
                      violation_code == "CC942" | 
                      violation_code == "CC943" | 
                      violation_code == "CC948" | 
                      violation_code == "CC956" | 
                      violation_code == "CC921"),
    cc941_pct = round((total_941/total_940)*100, 2),
    other_pct = round((total_other/total_940)*100, 2)
  ) |> 
  ungroup() |>
  arrange(off_floor_mo)

codes_by_pct 
```

##### 3.3.2.2 Plotting

```{r}
codes_by_pct_plot <- ggplot(
  codes_by_pct,
  aes(x = off_floor_mo)) +
  geom_line(aes(y = cc941_pct,
                color = "CC941")) +
  geom_point(aes(y = cc941_pct,
                color = "CC941"),
             size = 0.4) +
  geom_line(aes(y = other_pct,
                color = "All others")) +
  geom_point(aes(y = other_pct,
                color = "All others"),
             size = 0.4) +
  geom_vline(xintercept = ymd("2024-01-01"),
             linetype = "dashed", 
             color = "grey70") +
  geom_vline(xintercept = ymd("2025-07-01"),
             linetype = "dashed", 
             color = "grey70") +
  labs(
    title = str_wrap("CC941 and all 'other' codes as percentages of CC940"),
    subtitle = str_wrap("As 941's percentage decreases, 'other' violation code percentages increase. Starting around July 2025, when Whitmire introduced new ordinances, there became less CC941 citations relative to CC940 numbers, while new ordinance citations increased."),
    caption = "By Layla Dajani",
    x = "Year",
    y = "Percent of CC940"
      ) +
theme_minimal()

codes_by_pct_plot |> 
  ggplotly() #@ add new column with descriptors 1 word
```

##### 3.3.2.3 Data takeaways

@write

### 3.3 Ordinance types over time

#### 3.3.1 Making object

Grouping by ordinance type and plotting over time. @ fix x axis stops

```{r}
types_over_time <- unhoused_distinct |>
  filter(offense_date >= "2023-07-01") |> 
  group_by(off_floor_mo, 
           ordinance_type) |>
  summarise(citation_count = n(), .groups = "drop") |>
  arrange(off_floor_mo)

types_over_time
```

#### 3.3.2 Plotting - for summary

```{r}
types_over_time_plot <- ggplot(
  types_over_time,
  aes(x = off_floor_mo, 
      y = citation_count, 
      fill = ordinance_type
      )) +
  geom_col(position = "stack") +
  geom_vline(xintercept = ymd("2024-01-01"),
             linetype = "dashed", 
             color = "grey40") +
  geom_vline(xintercept = ymd("2025-07-01"),
             linetype = "dashed", 
             color = "grey40") +
  labs(
    x = "Date",
    y = "Number of citations per month",
    fill = "Violation type",
    title = "Monthly citations by violation type from Jul. 2023 to Sept. 2025"
      ) +
  theme_minimal()

ggsave("figures/types-over-time-plot.png")

types_over_time_plot |> 
  ggplotly()
```

#### 3.3.3 Data takeaways

October 2025 had highest sidewalk obstruction citations per month in this data set. The 146 citations made up around 30% of the total citations from that month.

## 4. Judgment and dispositions

### 4.1 Most common judgment

What is the most common specific judgment?

```{r}
unhoused |> # not using unhoused_distinct here to show the true judgment stats. even if there are repeated cases, those repeats each received their own judgment in court.
  group_by(judgment) |> 
  summarise(judgment_count = n()) |> 
  arrange(desc(judgment_count))
```

What is the most common refined judgment?

```{r}
unhoused |>
  group_by(judgment_refined) |> 
  summarise(judgment_count = n()) |>
  arrange(desc(judgment_count))
```

#### 4.1.1 Data takeaways

* Judgment unrefined
  + The most common judgment is "CASE DISMISSED - OFFICER NOT PRESENT" with 8,439	counts, followed by "CREDIT TIME SERVED" with 5,440 counts and "FAILURE TO APPEAR WARRANT (FTA)" with 5,114 counts.
* Judgment refined
  + The most common judgment is "CASE DISMISSED" with 11,693 counts, followed by "CREDIT TIME SERVED" with 5,440 counts and "FAILURE TO APPEAR WARRANT (FTA)" with 5,144 counts.

### 4.2 Most common disposition

What is the most common disposition?

```{r}
unhoused |> 
  group_by(disposition) |> 
  summarise(disposition_count = n()) |> 
  arrange(desc(disposition_count))
```

#### 4.2.1 Data takeaways

* Since "UNDISPOSED" means the case still has not been decided, of the cases that have received a disposition, about 59% were dismissed at trial.

### 4.3 Yearly guilty cases

What is the year with the most guilty dispositions?

#### 4.3.1 Creating object

```{r}
yearly_guilty <- unhoused |> 
  filter(disposition == "GUILTY TRIAL BY JUDGE") |> 
  group_by(off_floor_yr) |> 
  summarize(yearly_guilty = n()) |> 
  arrange(off_floor_yr) 

yearly_guilty
```

#### 4.3.2 Plotting

```{r}
yearly_guilty_plot <- ggplot(
  yearly_guilty,
  aes(x = off_floor_yr, 
      y = yearly_guilty
      )) +
  geom_col(fill = "#4E79A7") +
  geom_vline(xintercept = ymd("2024-01-01"),
             linetype = "dashed", 
             color = "grey40") +
  labs(
    x = "Year",
    y = "Number of guilty judgments per month",
    title = "Guilty judgments by year from 2016 to 2025",
    subtitle = "The dotted line indicates when Whitmire's term began."
      ) +
  theme_minimal()

yearly_guilty_plot |> 
  ggplotly()
```

#### 4.3.1 Data takeaways

* 2018 is the year with the lowest number of guilty dispositions at 50 dispositions.
* 2016 is the year with the highest number of guilty dispositions at 1951 dispositions, followed by 2021 with 931.

### 4.4 Highest guilty month

Of all the individual months each year, what is the month with the most guilty dispositions percentage relative to all dispositions that month?

```{r}
guilty_months_pct <- unhoused |> 
  #filter(judgment_date >= "2023-07-01") |> 
  group_by(jud_yr,
           jud_floor_mo,
           whitmire_judgment) |> 
  summarise(guilty = sum(disposition == "GUILTY TRIAL BY JUDGE"),
            total = n(),
            guilty_pct = round((guilty/total) * 100, 2)
            ) |> 
  arrange(jud_floor_mo)

guilty_months_pct
```

#### 4.4.1 Plotting - for summary

@for summary

```{r}
guilty_months_pct_plot <- ggplot(
  guilty_months_pct, 
  aes(x = jud_floor_mo, 
      y = guilty_pct, 
      color = whitmire_judgment)) +
  geom_line(size = 0.9) +
  geom_point(size = 0.8) +
  geom_vline(xintercept = ymd("2024-01-01"),
             linetype = "dashed", 
             color = "grey40") +
  geom_vline(xintercept = ymd("2025-07-01"),
             linetype = "dashed", 
             color = "grey40") +
  scale_color_manual(
    values = c(
      "BEFORE WHITMIRE" = "#4E79A7", 
      "AFTER WHITMIRE"  = "#E15759")
    ) +
  labs(
    title = "Percentage of guilty dispositions per month",
    subtitle = str_wrap(""),
    x = "Date",
    y = "Percentage of guilty dispositions",
    color = ""
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("figures/guilty_months_pct_plot.png")

guilty_months_pct_plot |> 
  ggplotly()
```

#### 4.4.2 Data takeaways
@ write this
  
### 4.5 Avg monthly guilty

What is the average percentage of guilty dispositions for each month since 2016?

```{r}
avg_monthly_guilty <- unhoused |> 
  #filter(judgment_date < "2025-01-01") |> # 2025 is not a complete year so including it skews later months' results
  group_by(jud_yr,
           jud_mo_label) |>
  summarise(guilty = sum(disposition == "GUILTY TRIAL BY JUDGE"),
            total = n(),
            guilty_pct = (guilty/total) * 100
            ) |>
  group_by(jud_mo_label) |>
  summarise(avg_guilty_pct = round(mean(guilty_pct), 2)) |> 
  arrange(avg_guilty_pct)

avg_monthly_guilty
```

#### 4.5.2 Data takeaways

* Up to 2024
  + Looking at an average of every month from 2016 to 2024, the month with the lowest average percentage of guilty dispositions is September at 14.02%, followed by December at 14.12%.
  + The month with the highest average percentage of guilty dispositions is March at 24.86%, followed by January at 21.66%.
* Including 2025
  + Looking at an average of every month from January 2016 to Nov. 2025, the month with the lowest average percentage of guilty dispositions is November at 13.17%, followed by December at 12.42%.
  + The month with the highest is March with 23.2%, followed by June with 19.96%.

## 5. Top people

### 5.1 Distinct people

How many distinct people have received a citation?

```{r}
unhoused_distinct |> 
  summarise(people_count = n_distinct(refined_name))
```

#### 5.1.1 Data takeaway

* There are 4,682 distinct people who have received a citation since 2016. Note that this number is an estimate as the name cleaning process is not perfect.

### 5.2 Most citations

Who has received the most citations? 

```{r}
unhoused_distinct |> 
  group_by(refined_name) |> 
  summarise(citation_count = n(),
            first_citation = min(offense_date),
            last_citation = max(offense_date),
            ) |> 
  arrange(desc(citation_count)) |> 
  filter(citation_count >= 100)
```

#### 5.2.1 Data takeaways

* There are 26 people who have received at least 100 total citations. 
* The person with the most citations is Patricia Norris, who has received 794 total citations, all given between October 2017 and May 2025. 
* The person with the second most citations is Forest H Smith, who has received 351 total citations, all given between August 2020 and November 2025.

### 5.3 Most fines and dues

Who are the most heavily fined people? How much of those fines got dismissed? How much did they pay? How much do they still have due? Who has paid the highest amount of dues?

```{r}
individual_fines <- unhoused_distinct |>
  group_by(refined_name) |>
  summarise(total_fined = sum(amount_fined, na.rm = TRUE),
    total_fines_dismissed = sum(fines_dismissed, na.rm = TRUE),
    total_fines_paid = sum(fines_paid, na.rm = TRUE),
    total_fines_due = sum(fines_due, na.rm = TRUE)
  ) |>
  arrange(desc(total_fines_paid))

individual_fines
```

#### 5.3.1 Data takeaways

* Patricia Norris has been charged the highest total fines at around \$199,123. She hasn't paid any of her fines, but has had about \$149,118 dismissed, which leaves about \$50,006 that she still owes. 
* Forest H Smith is the second most fined person, with around \$86,223 total fines. He has also never paid any of his fines, but has had roughly \$9,584 dismissed, leaving around \$76,640 that he still owes.
* Kevin P Shakesnider is the person who has paid the highest amount of fines. He has paid about \$28,684 of his total \$45,125, and has had roughly \$10,916 dismissed. As of November 2025, he still owes about \$5,525.

## 6. Fines

1. How many distinct people are there who were fined?
2. How many people still owe fines to the city? How much is the city owed in total?
3. How many people have paid some/all of their fines?
4. How many people have had some/all of their fines dismissed?

### 6.1 Fined people

How many distinct people are there who were fined?

```{r}
individual_fines |> 
  filter(total_fined > 0) |> 
  summarise(fined_people = n(),
            amount_fined_total = sum(total_fined))
```

### 6.2 Owed fines

How many people still owe fines to the city? How much is the city owed in total?

```{r}
individual_fines |> 
  filter(total_fines_due > 0) |> 
  summarise(people_owing_fines = n(),
            owed_fines_total = sum(total_fines_due)
            )
```

### 6.3 Fines paid

#### 6.3.1 Some fines paid

How many people have paid at least some of their fines?

```{r}
individual_fines |> 
  filter(total_fines_paid > 0) |> 
  summarise(people_paid_some_fines = n(),
            paid_fines_total = sum(total_fines_paid) 
            )
```

#### 6.3.2 All fines paid - no dismissals

How many people have paid all of their fines with no help from dismissals?

```{r}
individual_fines |> 
  filter(total_fines_paid == total_fined) |> 
  summarise(people_paid_all_fines = n(),
            paid_fines_total = sum(total_fines_paid) 
            )
```

### 6.4 Fines dismissed

How many people have had some/all of their fines dismissed?

```{r}
individual_fines |> 
  filter(total_fines_dismissed > 0,
         total_fines_dismissed == total_fined) |> 
  summarise(individual_fines_dismissed = n()
            )
```

### 6.5 Data takeaways

1. There are about 4,456 distinct people who have received a fine with their citation since 2016, their cumulative total being $6,735,614.
2. Of the 4,456 distinct people who have received a fine with their citation, 2,972 of them still owe fines. The sum of total fines still owed to the city is \$3,013,654.	
3. Of the 4,456 distinct people who have received a fine with their citation, 1,225 of them have paid at least some of their fines. 785 of those people have paid off all their fines without any help from dismissals. The total amount of fines paid since 2016 is about \$1,486,821.	
4. Of the 4,456 distinct people who have received a fine with their citation, 1459 people have had at least some of their fines dismissed, and 699 of those 1,459 people had all of their fines dismissed.

## 7. Jail time

### 7.1 Credit time served

#### 7.1.1 Distinct jailed judgments

How many individual people have received at least one CREDIT TIME SERVED judgment? Who has received the most CREDIT TIME SERVED judgments?

```{r}
unhoused |> 
  group_by(refined_name) |> 
  summarise(credit_jail_judgments = sum(time_served == "yes")) |> 
  filter(credit_jail_judgments > 0) |> 
  arrange(desc(credit_jail_judgments))
```

#### 7.1.2 Distinct judgment dates

Who has received the most CREDIT TIME SERVED judgments on distinct judgment dates? 

**Explanation:** oftentimes, courts rule on several citations for the same person on one day. For example, Kevin Shakesnider didn't receive all 136 of his CREDIT TIME SERVED judgments on 136 separate days, he was assigned the same judgment date for several of his citations.

```{r}
unhoused |> 
  filter(time_served == "yes") |> 
  distinct(refined_name, judgment_date) |> 
  group_by(refined_name) |> 
  summarise(credit_jail_judgment_days = n()) |> 
  #filter(credit_jail_judgment_days > 1) |> 
  arrange(desc(credit_jail_judgment_days))
```

#### 7.1.3 Data takeaways

* Of the 4,598 people who have received a citation since 2016, all 4,598 of them received a "CREDIT TIME SERVED" judgment at least once. Of those 4,598, 841 people have received more than one "CREDIT TIME SERVED" judgment.
* The person with the highest credit_jail_judgments is Kevin P Shakesnider, who has received 136 individual "CREDIT TIME SERVED" judgments. Those 136 judgments were given across 22 distinct judgment dates.
* The person with the second highest credit_jail_judgments count is BOBBY L ANDERSON JR, who has received 104 individual "CREDIT TIME SERVED" judgments. Those 104 judgments were given across 7 distinct judgment dates.
* Note that although Patricia Norris has received the highest citation count, she has never received a "CREDIT TIME SERVED" judgment.

### 7.2 Motion granted inmate

#### 7.2.1 Individual jailed judgments

How many individual people have received at least one MOTION GRANTED INMATE (FULL CREDIT) judgment? Who has received the most MOTION GRANTED INMATE (FULL CREDIT) judgments?

```{r}
unhoused |> 
  group_by(refined_name) |> 
  summarise(motion_jail_judgments = sum(judgment == "MOTION GRANTED INMATE (FULL CREDIT)")) |> 
  filter(motion_jail_judgments > 0) |> 
  arrange(desc(motion_jail_judgments))
```

#### 7.2.2 Distinct judgment dates

Who has received the most MOTION GRANTED INMATE (FULL CREDIT) judgments on distinct judgment dates? 

**Explanation:** oftentimes, courts decide on several citations for the same person on one day. For example, Kevin Shakesnider didn't receive all 136 of his CREDIT TIME SERVED judgments on 136 separate days, the courts gave him the same judgment date for several of his citations.

```{r}
unhoused |> 
  filter(judgment == "MOTION GRANTED INMATE (FULL CREDIT)") |> 
  distinct(refined_name, judgment_date) |> 
  group_by(refined_name) |> 
  summarise(motion_jail_judgment_days = n()) |> 
  #filter(motion_jail_judgment_days > 1) |> 
  arrange(desc(motion_jail_judgment_days))
```

#### 7.2.3 Data takeaway

* Of the 4,598 people who have received a citation since 2016, 35 of them received a "MOTION GRANTED INMATE (FULL CREDIT)" judgment at least once. Of those 35, only one person has received more than one "MOTION GRANTED INMATE (FULL CREDIT)" judgment, and that is Robert F Living.
* The person with the highest motion_jail_judgments is Melissa Clark, who has received 12 individual "MOTION GRANTED INMATE (FULL CREDIT)" judgments. Those 12 judgments were given one distinct judgment date.
* Robert F Living is the only person with more than one "MOTION GRANTED INMATE (FULL CREDIT)" judgment day, at two judgment days. Between those two days, he received a total of four "MOTION GRANTED INMATE (FULL CREDIT)" judgments.

### 7.3 Jail judgments over time

What's the month with the most jail judgments from 2016 to present?

```{r}
unhoused |> 
  filter(judgment == "CREDIT TIME SERVED") |>
  group_by(jud_yr,
           jud_mo_label,
           jud_mo) |> 
  summarise(
    yr_mo_jail = n()) |> 
  arrange(desc(yr_mo_jail))
  #arrange(jud_yr, jud_mo)
```

#### 7.3.1 Data takeaways

* SINCE 2016:
  + The month with the highest "CREDIT TIME SERVED" judgments since 2016 is Jun. 2016 with 185 judgments. 
* SINCE JULY 2023:
  + The month with the highest "CREDIT TIME SERVED" judgments since July 2023 is Aug. 2023 with 57 citations.

## 8. Individual people analysis

### 8.1. Kevin P Shakesnider

Shakesnider data takeaways refresher:
* **Highest fines:** Kevin P Shakesnider is the person who has paid the highest amount of fines. He has paid about \$28,684 of his total \$44,453, and has had roughly \$10,438 dismissed. As of October 2025, he still owes about \$5,331.
* **Highest "CREDIT TIME SERVED" judgments:** The person with the highest credit_jail_judgments is Kevin P Shakesnider, who has received 136 individual "CREDIT TIME SERVED" judgments. Those 136 judgments were given across 22 distinct judgment dates.

#### 8.1.1 Highest "CREDIT TIME SERVED" judgments

When did Kevin P Shakesnider receive the highest jail citations in one judgment? What is the overall date range?

```{r}
unhoused_distinct |> 
  filter(refined_name == "SHAKESNIDER, KEVIN P") |>
  group_by(judgment_date, violation_code) |>
 summarise(jail_count = n())
```

##### 8.1.1.1 Data takeaways

All 21 of his judgment dates fall between Jan. 1 and Dec. 22 2016. On June 23, he received a "CREDIT TIME SERVED" judgment for 16 citations, his highest in one single day.

#### 8.1.2 Plotting

Plotting citations and violation type.

##### 8.1.2.1 Preparing object

```{r}
shakesnider_jailed <- unhoused |> 
  filter(refined_name == "SHAKESNIDER, KEVIN P") |>
  group_by(judgment_date, violation_code) |>
 summarise(jail_count = n())

shakesnider_jailed
```

### 8.2 Patricia Norris

#### 8.2.1 Highest citation count

Making Norris object.

```{r}
norris <- unhoused |> 
  filter(refined_name == "NORRIS, PATRICIA")

norris
```

#### 8.2.2 Plotting

##### 8.2.2.1 Preparing to plot

Checking for what kinds of violation codes Norris has been given.

```{r}
norris |> 
  distinct(violation_code)
```

Making the object.

```{r}
norris_citations <- norris |> 
  group_by(offense_date) |> 
  summarise(citation_count = n()) |>
  arrange(offense_date)

norris_citations
```

##### 8.2.2.2 Plotting - for summary

```{r}
norris_citations_plot <- ggplot(
  norris_citations, 
       aes(x = offense_date, 
           y = citation_count)) +
  geom_point(alpha = 1 / 3) +
  labs(
    title = "Patricia Norris Citation Count by Offense Date",
    x = "Offense Date",
    y = "Citation Count"
  ) +
  theme_minimal()

ggsave("figures/norris_citations_plot.png")

norris_citations_plot |> ggplotly()
```

###### 8.2.2.1.1 Data takeaways

* Each one of these points is a day, so the higher the citation count, the more citations she received in a single day. Almost every time Patricia Norris has been given a citation, she has also been given at least one other citation. She was also hit with citations almost consistently from January 2020 to May 2025.

## 9. Mapping analysis

### 9.1 Preparing mapping object

#### 9.1.1 Importing district data

Importing Houston district data as an sf object.

```{r}
hou_districts <- st_read(
  "https://mycity2.houstontx.gov/pubgis02/rest/services/HoustonMap/Administrative_Boundary/MapServer/2/query?where=1=1&outFields=*&f=geojson"
)

hou_districts
```

#### 9.1.2 Simple feature

Turning the addresses in a Simple Feature Collection.

```{r}
pts <- st_as_sf(
  unhoused_distinct,
  coords = c("geocodio_longitude", "geocodio_latitude"),
  crs = 4326
)

pts |> glimpse()
```

#### 9.1.3 Fixing invalid district lines

```{r}
clean_districts <- st_make_valid(hou_districts) # this function goes through every polygon to fix problems like duplicate points and self intersections, which i was having trouble with.
```

```{r}
cleaner_districts <- st_transform(clean_districts, 4326)
```

#### 9.1.4 Citation to district

Assigning each citation to a district.

```{r}
districting_citations <- st_join(pts, cleaner_districts, join = st_within)

districting_citations
```

Cleaning names after joining:

```{r}
districting_citations_clean <- districting_citations |> 
  clean_names()

districting_citations_clean |> glimpse()
```

### 9.2 Citations by district

Which district has received the most citations from 2016 to Sept. 2025? 

#### 9.2.1 Counting citation per district

Grouping data by district, preparing to map.

```{r}
district_counts <- districting_citations_clean |>
  st_drop_geometry() |> # this removes the geometry so we can summarize citation points, but i will add geometry back to map @ double check missing zones
  count(district, name = "citation_count")

district_counts
```

#### 9.2.2 Rejoining

Cleaning cleaner_district names before joining.

```{r}
cleaner_names_districts <- clean_names(cleaner_districts)

cleaner_names_districts |> glimpse()
```

Joining

```{r}
unhoused_to_map <- cleaner_names_districts |>
  left_join(district_counts, by = "district")

unhoused_to_map
```

### 9.6 Plotting

#### 9.6.1 General ggplot

Plotting a stagnant heat map to show each district's citations.

```{r}
ggplot(unhoused_to_map) +
  geom_sf(aes(fill = citation_count), color = "white", size = 0.3) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    title = "Citations given to unhoused people by Houston City Council district",
    caption = "By Layla Dajani",
    fill = "Citations",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()
```

#### 9.6.1 General mapview

Plotting an interactive heat map to show each district's citations.

```{r}
mapview(
  unhoused_to_map,
  zcol = "citation_count",
  popup = c("district", "citation_count"),
  layer.name = "Citations"
) #@ by percentage increase two years before and after
```







